### 2025-01-29 products/manuscript/Manuscript.qmd
Part 1 Commit 1 - research question, hypothesis, methods, data

---
title: "Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022"
author: "Rayleen Lewis"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(openxlsx)
```

**Authors**

* Rayleen Lewis$^{1}$ (ORCID: 0000-0000-1234-5678)

**Author affiliations**

1. College of Public Health, University of Georgia, Athens, GA, USA.

$\land$ Corresponding author: some@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the
author's own and don't reflect their employer or University.

{{< pagebreak >}}

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

## Questions/Hypotheses to be addressed
The objective of this work is to determine if HPV vaccination coverage changed between 2019 and 2022 among countries offering HPV vaccination as of 2019, and if so, are these changes correlated with COVID-19 burden in 2020-2021. I hypothesize that HPV vaccination coverage was lower in 2022 than 2019 among countries with high burden of COVID-19 in 2020-2021 and comparable in countries with a low COVID-19 burden.

{{< pagebreak >}}

# Methods 
Briefly, all datasets will be reduced to include only countries with vaccination coverage estimates in both 2019 and 2022. Average coverage will be compared in 2022 to 2019. Multivariable linear regression modeling the difference in coverage between the two years as the outcome will be conducting using measures of COVID-19 burden, HPV vaccination program characteristics, country income information, and country-level demographics. Measure(s) of COVID-19 burden will be considered the explanatory variable of interest. Other characteristics will be considered for adjustment.

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard). HPV Dashboard include country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccine schedule, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (target males and females), WHO region, and year-specific coverage from 2010 to 2023.

Country-specific weekly COVID-19 case and death counts are available from starting in 1/4/2020 through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o). In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID-19 data.

Additional country-level covariates were collected from other publicly available data sources. Information on country income level by year is available through the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups#:~:text=For%20the%20current%202025%20fiscal,those%20with%20a%20GNI%20per). Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)

All data sources are publicly available for download and were downloaded on January 26-27, 2025.

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._
```{r, echo=FALSE}
#world_bank <- read.xlsx("world-bank-country-income.xlsx", sheet = "Country Analytical History")

```

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

### 2025-01-30 products/manuscript/Manuscript.qmd
Part 1 Commit 2 - added data details

@@
 library(openxlsx)
+library(naniar)
 ```
@@
 
-$\land$ Corresponding author: some@email.com
+$\land$ Corresponding author: rml20258@uga.edu
 
@@
 ## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+Information on the dat sources can be found in the Data Acquisition and Data Importing and Cleaning sections below.
+Briefly, all datasets will be reduced to include only countries with vaccination coverage estimates in both 2019 and 2022.
+Average coverage will be compared in 2022 to 2019.
+Multivariable linear regression modeling the difference in coverage between the two years as the outcome will be conducting using measures of COVID-19 burden, HPV vaccination program characteristics, country income information, and country-level demographics.
+Measure(s) of COVID-19 burden will be considered the explanatory variable of interest.
+Other characteristics will be considered for adjustment.
 
@@
 Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).
+Briefly, coverage data reflect administrative and official HPV vaccination coverage reported annually through the WHO/UNICEF Joint Reporting Form on Immunization.
 HPV Dashboard include country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccine schedule, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (target males and females), WHO region, and year-specific coverage from 2010 to 2023.
@@
 All materials, including files that contain code, should be commented well so everyone can follow along._
+Each data source must be imported. Data sources can be merged based on country. Each dataset will be restricted to countries with vaccination coverage estimates in 2019 and 2022.
+
+#### HPV coverage and vaccination program data
+There are two data files to be imported for the HPV vaccination data, one file contains yearly coverage estimates and the other contains information about the HPV vaccination program.
+
+##### HPV vaccination coverage
+A total of 196 countries are included in the data set.
+Coverage estimates are available from 2010 to 2023, but most countries are missing on some years of data.
+Data manipulation will be needed to restrict to the years of interest and convert data form long to wide.
+
+```{r}
+#Importing yearly HPV vaccination coverage
+hpv_coverage_raw <- read.xlsx(here("data","raw-data","Human Papillomavirus (HPV) vaccination coverage 2025-21-01 11-57 UTC.xlsx"), sheet = "Sheet1", colNames = TRUE)
+
+#There are missing data, data will need to be converted from long to wide format to get a better sense of complete data
+gg_miss_var(hpv_coverage_raw)
+#Gives total number of countries in the dataset
+table(hpv_coverage_raw$YEAR)
+
+```
+
+##### HPV vaccination program characteristics
+A total of 196 countries are included in the data set.
+About one third of countries are missing year of introduction, and a few are missing country income.
+Other characteristics (e.g., delivery strategy) appear complete.
+```{r}
+#Importing HPV vaccination program characteristics
+hpv_program_raw <- read.csv(here("data","raw-data","who-dashboard-1dosecoverage-1-27-25.csv"), header = TRUE)
+
+#There are missing data, year of introduction is missing for about a third of countries
+gg_miss_var(hpv_program_raw)
+#Gives total number of countries in the dataset
+nrow(hpv_program_raw)
+```
+
+#### COVID19 data 
+This data source includes weekly COVD case count data on 239 countries.
+The data set includes 440640 rows, but about half have missing data on new case counts.
+Completeness of data will have to be reviewed once the data are restricted to countries with vaccination coverage estimates in the years of interest.
+```{r}
+#Importing HPV vaccination program characteristics
+covid_raw <- read.csv(here("data","raw-data","WHO-COVID-19-global-daily-data.csv"), header = TRUE)
+
+#There are missing data, year of introduction is missing for about a third of countries
+gg_miss_var(covid_raw)
+#Gives total number of countries in the dataset
+nrow(covid_raw)
+#Getting the number of countires - 239
+countries <- table(covid_raw$Country_code)
+dim(countries)
+```
+
+#### United Nations data 
+There are 238 countries with yearly data from 1950 to 2023.
+Data appear mostly complete (ISO2 Alpha-code is missing for some rows but this is missing for all World and Region specific rows since this is a code to indicate country).
+Data will be restricted to the two years of interest during the next part of the project.
+```{r}
+
+u_n_raw <- read.xlsx(here("data","raw-data","WPP2024_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT.xlsx"), sheet = "Estimates")
+header_row <- 10
+column_names_un <- as.character(unlist(u_n_raw[header_row, ]))
+
+# Read the data again, skipping the rows before the header
+u_n <- read.xlsx(here("data","raw-data","WPP2024_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT.xlsx"), sheet = "Estimates", startRow = 18, colNames = FALSE)
+colnames(u_n) <- column_names_un
+
+#Getting number of countries
+un_countries <- table(u_n$`ISO3 Alpha-code`)
+dim(un_countries)
+#238 countries
+
+#Getting number of years
+table(u_n$Year)
+#Year ranges from 1950 to 2023
+
+#There are missing data, year of introduction is missing for about a third of countries
+gg_miss_var(u_n)
+```
+
+#### World Bank data 
+There are 218 countries with yearly data from 1987 to 2023.
+Some data are missing, but this is mostly complete.
+Data will be restricted to the two years of interest during the next part of the project.
 ```{r, echo=FALSE}
-#world_bank <- read.xlsx("world-bank-country-income.xlsx", sheet = "Country Analytical History")
+
+world_bank_raw <- read.xlsx(here("data","raw-data","world-bank-country-income.xlsx"), sheet = "Country Analytical History")
+header_row <- 4
+column_names <- as.character(unlist(world_bank_raw[header_row, ]))
+
+# Read the data again, skipping the rows before the header
+world_bank <- read.xlsx(here("data","raw-data","world-bank-country-income.xlsx"), sheet = "Country Analytical History", startRow = 12, colNames = FALSE)
+colnames(world_bank) <- column_names
+colnames(world_bank)[c(1, 2)] <- c("country_code", "country")
+
+world_bank <- as_tibble(world_bank[1:218,]) 
+tail(world_bank)
+#218 countries
 

### 2025-02-20 Code chunks for reference.qmd
Part 2 Commit 1 - finished data cleaning, added eda, updated READMEs, deleted unneeded files

--- 2025-01-29 products/manuscript/Manuscript.qmd
@@
 ---
-title: "Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022"
-author: "Rayleen Lewis"
-format:
-  docx:
-    toc: false
-    number-sections: true
-    highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+title: "Code chunks for review"
+format: html
+editor: visual
 ---
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
+## Quarto
 
-```{r, echo=FALSE, message=FALSE}
-# load a few R packages
-library(here)
-library(knitr)
-library(tidyverse)
-library(openxlsx)
+Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
+
+## Running Code
+
+When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
+
+```{r}
+1 + 1
 ```
 
-**Authors**
+You can add options to executable code like this
 
-* Rayleen Lewis$^{1}$ (ORCID: 0000-0000-1234-5678)
+```{r}
+#| echo: false
+2 * 2
+```
 
-**Author affiliations**
-
-1. College of Public Health, University of Georgia, Athens, GA, USA.
-
-$\land$ Corresponding author: some@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the
-author's own and don't reflect their employer or University.
-
-{{< pagebreak >}}
-
-# Summary/Abstract
-_Write a summary of your project._
-
-{{< pagebreak >}}
-
-# Introduction
-
-## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
-
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
-
-## Questions/Hypotheses to be addressed
-The objective of this work is to determine if HPV vaccination coverage changed between 2019 and 2022 among countries offering HPV vaccination as of 2019, and if so, are these changes correlated with COVID-19 burden in 2020-2021.
-I hypothesize that HPV vaccination coverage was lower in 2022 than 2019 among countries with high burden of COVID-19 in 2020-2021 and comparable in countries with a low COVID-19 burden.
-
-{{< pagebreak >}}
-
-# Methods 
-Briefly, all datasets will be reduced to include only countries with vaccination coverage estimates in both 2019 and 2022.
-Average coverage will be compared in 2022 to 2019.
-Multivariable linear regression modeling the difference in coverage between the two years as the outcome will be conducting using measures of COVID-19 burden, HPV vaccination program characteristics, country income information, and country-level demographics.
-Measure(s) of COVID-19 burden will be considered the explanatory variable of interest.
-Other characteristics will be considered for adjustment.
+The `echo: false` option disables the printing of code (only output is displayed).
 
@@
 ```
-
-## Data aquisition
-Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).
-HPV Dashboard include country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccine schedule, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (target males and females), WHO region, and year-specific coverage from 2010 to 2023.
-
-Country-specific weekly COVID-19 case and death counts are available from starting in 1/4/2020 through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o).
-In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID-19 data.
-
-Additional country-level covariates were collected from other publicly available data sources.
-Information on country income level by year is available through the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups#:~:text=For%20the%20current%202025%20fiscal,those%20with%20a%20GNI%20per).
-Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)
-
-All data sources are publicly available for download and were downloaded on January 26-27, 2025.
-
-## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
-```{r, echo=FALSE}
-#world_bank <- read.xlsx("world-bank-country-income.xlsx", sheet = "Country Analytical History")
-
-```
-
-## Statistical analysis
-_Explain anything related to your statistical analyses._
-
-{{< pagebreak >}}
 
@@
 
+## Exploratory/Descriptive analysis
+
+_Use a combination of text/tables/figures to explore and describe your data.
+Show the most important descriptive results here.
+Additional ones should go in the supplement.
+Even more can be in the R and Quarto files that are part of your project._
+
+@tbl-summarytable shows a summary of the data.
+
+Note the loading of the data providing a **relative** path using the `../../` notation.
+(Two dots means a folder up).
+You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
+You can also use the `here` R package to create paths.
+See examples of that below.
+I generally recommend the `here` package.
+
+```{r}
+#| label: tbl-summarytable
+#| tbl-cap: "Data summary table."
+#| echo: FALSE
+resulttable=readRDS("../../results/tables/summarytable.rds")
+knitr::kable(resulttable)
+```
+
 {{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
-
-## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
 
@@
 Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+
+Additional country-level covariates were collected from other publicly available data sources.
+Information on country income level by year is available through the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups#:~:text=For%20the%20current%202025%20fiscal,those%20with%20a%20GNI%20per).
+Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)

### 2025-02-20 code/eda-code/eda.qmd
Part 2 Commit 1 - finished data cleaning, added eda, updated READMEs, deleted unneeded files

---
title: "Exploratory analysis script"
date: "2024-02-07"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(gt)
library(gtsummary)
```

Loading the project data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","project_data.rds")
#load data
project_data <- readRDS(data_location)
```

# Data exploration through tables

The code below calculates the distribution of each categorical variable and basic statistics for continuous variables. 
```{r}
str(project_data)

#Getting distributions for categorical variables
freq_table <- proc_freq(project_data, tables = c("WHO_REGION","WB_INCOME","HPV_NATIONAL_SCHEDULE","HPV_YEAR_INTRODUCTION","HPV_PRIM_DELIV_STRATEGY","HPV_SEX","doses_rec"), options = "missing", output = out)
freqtable_file = here("results", "tables", "freq_table_results.rds")
saveRDS(freq_table, file = freqtable_file)

project_data %>% dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>% tbl_summary()

#Getting statistics for continuous variables
means_table <- proc_means(project_data, var = c("total_pop","pop_density","median_age","pop_growth_rate","fertility_rate","death_rate","life_expect","migration","Cumulative_case_rate","Cumulative_death_rate","HPV_YEAR_INTRODUCTION","coverage_diff"), stats = c("N","NMISS","MEAN","STD","MEDIAN","Q1","Q3","MIN","MAX"), options = v(maxdec = 2), output = out)
meanstable_file = here("results", "tables", "means_table_results.rds")
saveRDS(means_table, file = meanstable_file)

```

Next, I'll look at the continuous variables using histograms;
```{r}
project_data %>%
  ggplot(aes(total_pop)) + geom_histogram()

project_data %>%
  ggplot(aes(pop_density)) + geom_histogram()

project_data %>%
  ggplot(aes(median_age)) + geom_histogram()

project_data %>%
  ggplot(aes(pop_growth_rate)) + geom_histogram()

project_data %>%
  ggplot(aes(fertility_rate)) + geom_histogram()

project_data %>%
  ggplot(aes(death_rate)) + geom_histogram()

project_data %>%
  ggplot(aes(life_expect)) + geom_histogram()

project_data %>%
  ggplot(aes(migration)) + geom_histogram()

project_data %>%
  ggplot(aes(Cumulative_case_rate)) + geom_histogram()

project_data %>%
  ggplot(aes(Cumulative_death_rate)) + geom_histogram()

project_data %>%
  ggplot(aes(HPV_YEAR_INTRODUCTION)) + geom_histogram()

```

Next step is to look at the outcome (difference in coverage between 2022 and 2019).
```{r}
#histogram of differnce in coverage
project_data %>%
  ggplot(aes(coverage_diff)) + geom_histogram()

```
Final exploratory is to look at bivariate associations between each of the predictors and the outcome through figures.
```{r}
#Violin plots of the outcome by each categorical variable
project_data %>%
  ggplot(aes(coverage_diff, WHO_REGION)) + geom_violin()

project_data %>%
  ggplot(aes(coverage_diff, WB_INCOME)) + geom_violin()

project_data %>%
  ggplot(aes(coverage_diff, HPV_NATIONAL_SCHEDULE)) + geom_violin()

project_data %>%
  ggplot(aes(coverage_diff, HPV_PRIM_DELIV_STRATEGY)) + geom_violin()

project_data %>%
  ggplot(aes(coverage_diff, HPV_SEX)) + geom_violin()

project_data %>%
  ggplot(aes(coverage_diff, as.factor(doses_rec))) + geom_violin()

#Scatterplots with the outcome and each continuous variable
project_data %>%
  ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth()
#Taking a closer look at the data that is bunched up less than <100,000
project_data %>%
  filter(total_pop < 100000) %>%
  ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth()
project_data %>% filter(pop_density<1000) %>%
  ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, median_age)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, fertility_rate)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, death_rate)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, life_expect)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, migration)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, Cumulative_case_rate)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, Cumulative_death_rate)) + geom_point() + geom_smooth()

project_data %>%
  ggplot(aes(coverage_diff, HPV_YEAR_INTRODUCTION)) + geom_point() + geom_smooth()

```
There's pretty substantial variability in the outcome. There don't appear to be any strong relationships between the covariates and the outcome, and linear relationships are likely not great fits. For most associations, a second or third order term may be more appropriate for logistic regression modeling.

Based on the visual inspections and apriori knowledge, the following characteristics were considered for multivariable modeling: WHO region, country income level, year of HPV vaccination, primary delivery strategy, gender neutral vacciantion, number of recommended doses, median age, fertility rate, death rate, life expectancy, migration, cumulative case rate, cumulative death rate.

### 2025-02-20 code/eda-code/eda_for_manuscript.qmd
Part 2 Commit 1 - finished data cleaning, added eda, updated READMEs, deleted unneeded files

---
title: "edo_for_manuscript"
format: html
---

# Overview
All exploratory figures and tables presented in the manuscript are coded in this code file.

#### Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
library(gt)
library(gtsummary)
library(flextable)
library(gtExtras)
library(kableExtra)
```

Loading the project data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","project_data.rds")
#load data
project_data <- readRDS(data_location)
```

#### Outcome: Difference in HPV vaccination coverage between 2022 and 2019

```{r}
#histogram of differnce in coverage
outcome_hist <- ggplot(project_data, aes(x = coverage_diff)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(
    x = "Difference in HPV vaccination coverage from 2022 to 2019",
    y = "Count"
  )

outcome_hist_log <- here("results","figures","outcome_hist.png")
# Save the last plot created as a PNG
ggsave(outcome_hist_log, width = 8, height = 6, dpi = 300)

```

#### Creating "Table 1" to show the distribution of each characteristic
```{r}
  table1 <- project_data %>% dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>% #Reordering the variables into the order I want displayed in the table
  dplyr::select(coverage_2022, coverage_2019, coverage_diff, Cumulative_case_rate, Cumulative_death_rate, HPV_NATIONAL_SCHEDULE, HPV_YEAR_INTRODUCTION, HPV_PRIM_DELIV_STRATEGY, HPV_SEX, doses_rec, WHO_REGION, WB_INCOME, total_pop,  total_pop, pop_density, median_age, pop_growth_rate, fertility_rate, death_rate, life_expect, migration) %>%
  tbl_summary(
    type = all_continuous() ~ "continuous2",
    statistic = list(   #Setting the statistics to calculate
      all_continuous() ~ c(
      "{mean} ({sd})",
      "{median} ({p25}, {p75})"
    ),
      all_categorical() ~ "{n} / {N} ({p}%)"),
    digits = list(   #Setting number of decimal places for each variable
      coverage_2022 ~ 1,
      coverage_2019 ~ 1,
      coverage_diff ~ 1,
      Cumulative_case_rate ~ 1,
      Cumulative_death_rate ~ 1,
      HPV_YEAR_INTRODUCTION ~ function(x) sprintf("%.0f", x),
      total_pop ~ 1,
      pop_density ~ 1,
      median_age ~ 1,
      pop_growth_rate ~ 1,
      fertility_rate ~ 1,
      death_rate ~ 1,
      life_expect ~ 1,
      migration ~ 1
    )) %>%
  modify_header(
    stat_0 = "**n/N (%), mean (sd), or median (IQR),** N = {n}") %>%
  modify_footnote(all_stat_cols() ~ NA) %>%
  modify_table_styling(
    columns = label,    # left aligning "label" column
    align = "left"
  ) 
# modify_header(stat0~ "**n/N (%), mean (sd), or median (IQR),**\n N = {n}")

saveRDS(table1, here("results","tables","table1.rds"))

```

```{r}
#This is creating the combined table 1 that can be used to pull numbers from 
# Process the data and create a summary table using dplyr
word_list <- c("coverage_2022", "coverage_2019", "coverage_diff", "Cumulative_case_rate", 
    "Cumulative_death_rate", "HPV_NATIONAL_SCHEDULE", "HPV_YEAR_INTRODUCTION", 
    "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "doses_rec", "WHO_REGION", "WB_INCOME", 
    "total_pop", "pop_density", "median_age", "pop_growth_rate", "fertility_rate", 
    "death_rate", "life_expect", "migration")
table1_prelim <- project_data %>%
  dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
  dplyr::select(
    coverage_2022, coverage_2019, coverage_diff, Cumulative_case_rate, 
    Cumulative_death_rate, HPV_NATIONAL_SCHEDULE, HPV_YEAR_INTRODUCTION, 
    HPV_PRIM_DELIV_STRATEGY, HPV_SEX, doses_rec, WHO_REGION, WB_INCOME, 
    total_pop, pop_density, median_age, pop_growth_rate, fertility_rate, 
    death_rate, life_expect, migration
  ) %>%
  tbl_summary(
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = list(
      coverage_2022 ~ 1,
      coverage_2019 ~ 1,
      coverage_diff ~ 1,
      Cumulative_case_rate ~ 1,
      Cumulative_death_rate ~ 1,
      HPV_YEAR_INTRODUCTION ~ function(x) sprintf("%.0f", x),
      total_pop ~ 1,
      pop_density ~ 1,
      median_age ~ 1,
      pop_growth_rate ~ 1,
      fertility_rate ~ 1,
      death_rate ~ 1,
      life_expect ~ 1,
      migration ~ 1
    )
  ) %>%
  modify_header(
    stat_0 = "n/N (%), mean (sd), or median (IQR), N = {n}",
    label = "Characteristic"
  ) %>% as_tibble() %>%
  mutate(`n/N (%), mean (sd), or median (IQR), N = 100` = if_else(is.na(`n/N (%), mean (sd), or median (IQR), N = 100`), " ", `n/N (%), mean (sd), or median (IQR), N = 100`))

saveRDS(table1_prelim, here("results","tables","table1_tib.rds"))

table1 <- table1_prelim %>%
  gt() %>%
  tab_style(
    style = cell_text(weight = "bold"), # Bold text
    locations = cells_body(
      columns = Characteristic,
      rows = grepl(paste(word_list, collapse = "|"), Characteristic, ignore.case = TRUE)  # Apply bold if "mean" or "median" is NOT in the Variable column
    )
  )%>%
    #Making all column headers bld
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything()) 
  )

typeof(table1)
table1_df <- as_data_frame(table1_prelim) %>%as_flextable()
saveRDS(table1_df, here("results","tables","table1_df.rds"))

tab_1 <- gtsave(table1, filename = here("results","tables","table1.png"), expand = 10)

saveRDS(table1, here("results","tables","table1.rds"))
save_as_image(as_flex_table(table1), path = here("results","tables","table1.png"))
as_flextable(table1_display)

```

Because table rendering to Word appears broken right now, I'll split up the table and save the pieces as pictures, then insert the pictures into the manuscript document.
```{r}
#This is creating a "Table 1" for coverage and COVID burden
# Process the data and create a summary table using dplyr
coverage_covid <- c("coverage_2022", "coverage_2019", "coverage_diff", "Cumulative_case_rate", 
    "Cumulative_death_rate")
table1_cov_covid <- project_data %>%
  dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
  dplyr::select(
    coverage_2022, coverage_2019, coverage_diff, Cumulative_case_rate, Cumulative_death_rate
  ) %>%
  tbl_summary(
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = list(
      coverage_2022 ~ 1,
      coverage_2019 ~ 1,
      coverage_diff ~ 1,
      Cumulative_case_rate ~ 1,
      Cumulative_death_rate ~ 1
    )
  ) %>%
  modify_header(
    stat_0 = "Mean (sd) or median (IQR), N = {n}",
    label = "Characteristic"
  ) %>% as_tibble() %>%
  mutate(`Mean (sd) or median (IQR), N = 100` = if_else(is.na(`Mean (sd) or median (IQR), N = 100`), " ", `Mean (sd) or median (IQR), N = 100`)) %>%
  gt() %>%
  tab_style(
    style = cell_text(weight = "bold"), # Bold text
    locations = cells_body(
      columns = Characteristic,
      rows = grepl(paste(word_list, collapse = "|"), Characteristic, ignore.case = TRUE)  # Apply bold if "mean" or "median" is NOT in the Variable column
    )
  )%>%
    #Making all column headers bld
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything()) 
  ) %>%
  tab_options(
    table.font.size = px(10),        # Set font size
    data_row.padding = px(1),        # Reduce row padding
    column_labels.padding = px(2)    # Reduce padding in column headers
  )

gtsave(table1_cov_covid, filename = here("results","tables","table1_cov_covid.png"), expand = 15)

#########

#This is creating the "Table 1" for vaccination program variables
# Process the data and create a summary table using dplyr
vac_program <- c("HPV_NATIONAL_SCHEDULE", "HPV_YEAR_INTRODUCTION", 
    "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "doses_rec")
table1_vac_program <- project_data %>%
  dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
  dplyr::select(
    HPV_NATIONAL_SCHEDULE, HPV_YEAR_INTRODUCTION, 
    HPV_PRIM_DELIV_STRATEGY, HPV_SEX, doses_rec
  ) %>%
  tbl_summary(
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = list(
      HPV_YEAR_INTRODUCTION ~ function(x) sprintf("%.0f", x)
    )
  ) %>%
  modify_header(
    stat_0 = "n/N (%), mean (sd), or median (IQR), N = {n}",
    label = "Characteristic"
  ) %>% as_tibble() %>%
  mutate(`n/N (%), mean (sd), or median (IQR), N = 100` = if_else(is.na(`n/N (%), mean (sd), or median (IQR), N = 100`), " ", `n/N (%), mean (sd), or median (IQR), N = 100`)) %>% #Replacing NA with a space so that NA doesn't get printed in the table
  gt() %>%
  tab_style(
    style = cell_text(weight = "bold"), # Bold text
    locations = cells_body(
      columns = Characteristic,
      rows = grepl(paste(word_list, collapse = "|"), Characteristic, ignore.case = TRUE)  # Apply bold if "mean" or "median" is NOT in the Variable column
    )
  )%>%
    #Making all column headers bld
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything()) 
  ) %>%
  tab_options(
    table.font.size = px(10),        # Set font size
    data_row.padding = px(1),        # Reduce row padding
    column_labels.padding = px(2)    # Reduce padding in column headers
  )

gtsave(table1_vac_program, filename = here("results","tables","table1_vac_program.png"), expand = 15)

###########
#This is creating the combined table 1 that can be used to pull numbers from 
# Process the data and create a summary table using dplyr
demo_list <- c("WHO_REGION", "WB_INCOME", 
    "total_pop", "pop_density", "median_age", "pop_growth_rate", "fertility_rate", 
    "death_rate", "life_expect", "migration")
table1_demo <- project_data %>%
  dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
  dplyr::select(
    WHO_REGION, WB_INCOME, 
    total_pop, pop_density, median_age, pop_growth_rate, fertility_rate, 
    death_rate, life_expect, migration
  ) %>%
  tbl_summary(
    type = all_continuous() ~ "continuous2",
    statistic = list(
      all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} / {N} ({p}%)"
    ),
    digits = list(
      total_pop ~ 1,
      pop_density ~ 1,
      median_age ~ 1,
      pop_growth_rate ~ 1,
      fertility_rate ~ 1,
      death_rate ~ 1,
      life_expect ~ 1,
      migration ~ 1
    )
  ) %>%
  modify_header(
    stat_0 = "Mean (sd) or median (IQR), N = {n}",
    label = "Characteristic"
  ) %>% as_tibble() %>%
  mutate(`Mean (sd) or median (IQR), N = 100` = if_else(is.na(`Mean (sd) or median (IQR), N = 100`), " ", `Mean (sd) or median (IQR), N = 100`)) %>%
  gt() %>%
  tab_style(
    style = cell_text(weight = "bold"), # Bold text
    locations = cells_body(
      columns = Characteristic,
      rows = grepl(paste(word_list, collapse = "|"), Characteristic, ignore.case = TRUE)  # Apply bold if "mean" or "median" is NOT in the Variable column
    )
  )%>%
    #Making all column headers bld
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything()) 
  ) %>%
  tab_options(
    table.font.size = px(8),        # Set font size
    data_row.padding = px(1),        # Reduce row padding
    column_labels.padding = px(2)    # Reduce padding in column headers
  )

gtsave(table1_demo, filename = here("results","tables","table1_demo.png"), expand = 55)

```

#### Creating scatterplot of COVID cases and deaths by difference in coverage

```{r}

scatter_case_covdiff <- project_data %>%
  ggplot(aes(Cumulative_case_rate,coverage_diff)) + geom_point() + geom_smooth(method = lm) +
  labs(
    x = "Number of COVID cases in 2020-2022 per 100,000 persons",
    y = "Difference in HPV vaccination coverage between 2022 and 2019"
  ) 
scatter_case_covdiff_loc <- here("results","figures","scatter_case_covdiff.png")
# Save the last plot created as a PNG
ggsave(scatter_case_covdiff_loc, width = 8, height = 6, dpi = 300)

scatter_death_covdiff <- project_data %>%
  ggplot(aes(Cumulative_death_rate, coverage_diff)) + geom_point() + geom_smooth(method = lm) +
  labs(
    x = "Number of COVID deaths in 2020-2022 per 100,000 persons",
    y = "Difference in HPV vaccination coverage between 2022 and 2019"
  ) 
scatter_death_covdiff_loc <- here("results","figures","scatter_death_covdiff.png")
# Save the last plot created as a PNG
ggsave(scatter_death_covdiff_loc, width = 8, height = 6, dpi = 300)
```

#### Classfication tree

```{r}
# Load necessary libraries
library(rpart)
library(rpart.plot)

# Fit a regression tree model
reg_tree <- rpart(coverage_diff ~ total_pop + pop_density + median_age + pop_growth_rate + fertility_rate + death_rate + life_expect + migration + Cumulative_case_rate + Cumulative_death_rate + HPV_YEAR_INTRODUCTION + WHO_REGION + WB_INCOME + HPV_NATIONAL_SCHEDULE + HPV_YEAR_INTRODUCTION + HPV_PRIM_DELIV_STRATEGY + HPV_SEX + doses_rec, 
                  data = project_data, 
                  method = "anova")  # anova method for regression trees

# Visualize the tree
rpart.plot(reg_tree, type = 2, extra = 101, fallen.leaves = TRUE,
           main = "Regression Tree for Coverage Difference")

# Print a summary of the tree
summary(reg_tree)

```

### 2025-02-20 code/processing-code/processing-code.qmd
Part 2 Commit 1 - finished data cleaning, added eda, updated READMEs, deleted unneeded files

---
title: "Data Cleaning"
format: html
editor: visual
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(openxlsx)
library(naniar)
library(procs)
```

## Data import and cleaning

Each data source must be imported. Data sources can be merged based on country. Each dataset will be restricted to countries with vaccination coverage estimates in 2019 and 2022.

#### HPV coverage and vaccination program data
There are two data files to be imported for the HPV vaccination data, one file contains yearly coverage estimates and the other contains information about the HPV vaccination program.

##### HPV vaccination coverage
A total of 196 countries are included in the data set. Coverage estimates are available from 2010 to 2023, but most countries are missing on some years of data. Data manipulation will be needed to restrict to the years of interest and convert data form long to wide.

```{r}
#Importing yearly HPV vaccination coverage
hpv_coverage_raw <- read.xlsx(here("data","raw-data","Human Papillomavirus (HPV) vaccination coverage 2025-21-01 11-57 UTC.xlsx"), sheet = "Sheet1", colNames = TRUE)

#There are missing data, data will need to be converted from long to wide format to get a better sense of complete data
gg_miss_var(hpv_coverage_raw)
#Gives total number of countries in the dataset
table(hpv_coverage_raw$YEAR)

#Taking a look at the structure of the dataset
glimpse(hpv_coverage_raw)

table(hpv_coverage_raw$GROUP)
#Restricting data to years 2019 and 2022
hpv_coverage <- hpv_coverage_raw %>%
  filter(GROUP == "COUNTRIES") %>% #Deleting global estimates of country
  dplyr::select(CODE, NAME, YEAR, COVERAGE) %>% #Restricting to variables that will be used 
  filter(YEAR == 2022 | YEAR == 2019) %>% #Restricting to rows with year equal to 2019 or 2022
    #converting data from long to wide
    pivot_wider(
    names_from = YEAR, 
    values_from = COVERAGE,
    names_prefix = "coverage_"
  ) %>%
  rename(
    coverage_2019 = coverage_2019,
    coverage_2022 = coverage_2022
  ) %>%
  filter(!is.na(coverage_2019) & !is.na(coverage_2022))

summary(hpv_coverage)
head(hpv_coverage)

#Saving data to the processed data folder
saveRDS(hpv_coverage, file = here("data","processed-data","hpv_coverage.rds"))

```

##### HPV vaccination program characteristics
A total of 196 countries are included in the data set. About one third of countries are missing year of introduction, and a few are missing country income. Other characteristics (e.g., delivery strategy) appear complete.
```{r}
#Importing HPV vaccination program characteristics
hpv_program_raw <- read.csv(here("data","raw-data","who-dashboard-1dosecoverage-1-27-25.csv"), header = TRUE)

#There are missing data, year of introduction is missing for about a third of countries, more checks will be done later among the countries with coverage estimtes from 2019 and 2022 because this missingness is likely due to countries that havne't introduced HPV vaccination yet
gg_miss_var(hpv_program_raw)
#Gives total number of countries in the dataset
nrow(hpv_program_raw)

#Data cleaning
hpv_program <- hpv_program_raw %>%
  dplyr::select(-HPV1_COVERAGELASTYEAR,-HPVC_COVERAGELASTYEAR, -HPV_AGEADMINISTERED) %>%
  rename(CODE = ISO_3_CODE) %>% #renaming the ID variable, a unique country specific code, so this can be used to merge across datasets
  mutate(doses_rec = if_else(substr(HPV_INT_DOSES,1,1) == 1, 1,       if_else(substr(HPV_INT_DOSES,1,1) == 2, 2, NA)))

head(hpv_program)
summary(hpv_program)

#Saving data to the processed data folder
saveRDS(hpv_program, file = here("data","processed-data","hpv_program.rds"))

```

I am incrementally building the combined data set. As a first step, I am combining the HPV vaccination coverage estimates with the HPV vaccination program information and performing some basic data checks. Other combinations will be performed in the _Combining all datasets_ section below.
```{r}

combo <- hpv_coverage %>% left_join(hpv_program, by = join_by(CODE))
head(combo)
#Checking to see if any of the countries didn't match correctly by comparing the country names in each dataset
non_matches <- combo %>%
  filter(NAME != COUNTRYNAME)
print(non_matches)
#One country, Côte d'Ivoire, came up in this check, however, this only came up because the name of the country has some formatting issues in one of the datasets, this isn't a merging problem

#Data are 
proc_freq(combo, c("WHO_REGION", "WB_INCOME", "HPV_NATIONAL_SCHEDULE", "HPV_YEAR_INTRODUCTION", "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "HPV_INT_DOSES"), options = "missing")

#Data were very complete, at most variables had 1 country with missing data
#Reviewing the countries with missing data on any of the variables
combo_missing <- combo %>%
  filter(HPV_INT_DOSES=="" | WB_INCOME =="." )
print(combo_missing)
#Only 2 countries total were missing data on at least one variable, Bahamas and Palau, these countries will be retained since classification tree analysis is less affected by missing data

#Getting list of countries in the combo dataset to use for subsetting other datasets
countries <- combo %>% dplyr::select(CODE)

```

#### United Nations data 
There are 238 countries with yearly data from 1950 to 2023. Data appear mostly complete (ISO2 Alpha-code is missing for some rows but this is missing for all World and Region specific rows since this is a code to indicate country). Data will be restricted to the two years of interest during the next part of the project.
```{r}

u_n_raw <- read.xlsx(here("data","raw-data","WPP2024_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT.xlsx"), sheet = "Estimates")
header_row <- 10
column_names_un <- as.character(unlist(u_n_raw[header_row, ]))

# Read the data again, skipping the rows before the header
u_nations_full <- read.xlsx(here("data","raw-data","WPP2024_GEN_F01_DEMOGRAPHIC_INDICATORS_COMPACT.xlsx"), sheet = "Estimates", startRow = 18, colNames = FALSE)
colnames(u_nations_full) <- column_names_un

#Getting number of countries
u_nations_count <- table(u_nations_full$`ISO3 Alpha-code`)
dim(u_nations_count)
#238 countries

#Getting number of years
table(u_nations_full$Year)
#Year ranges from 1950 to 2023

#There are missing data, year of introduction is missing for about a third of countries
gg_miss_var(u_nations_full)

#Current dataset includes data at higher levels than just country, will check this again once data are restricted to countries of interest 
summary(u_nations_full)
tail(u_nations_full)
table(u_nations_full$`Type`)

#Note: this data includes both a 3 character and 2 character country-specific code, this is important because each other dataset includes one or the other, the United Nations data allows all data sets to be merged together
u_nations_full <- u_nations_full %>%
  rename(CODE = `ISO3 Alpha-code`) %>%
  rename(CODE2 = `ISO2 Alpha-code`)

#Restricting to countries of interest
u_nations_countries <- countries %>% left_join(u_nations_full, by = join_by(CODE))
summary(u_nations_countries)
head(u_nations_countries, n = 20)
#Only country level data is left
table(u_nations_countries$`Type`)

#Restricting to data columns of interest and renaming to easier to use names
u_nations <- u_nations_countries %>%
  filter(Year == 2019) %>%
  dplyr::select(CODE, CODE2,`Region, subregion, country or area *`, `Total Population, as of 1 July (thousands)`, `Population Density, as of 1 July (persons per square km)`, `Median Age, as of 1 July (years)`, `Population Growth Rate (percentage)`, `Total Fertility Rate (live births per woman)`, `Crude Death Rate (deaths per 1,000 population)`, `Life Expectancy at Birth, both sexes (years)`, `Net Migration Rate (per 1,000 population)`) %>%
  mutate(total_pop = as.numeric(`Total Population, as of 1 July (thousands)`), 
         pop_density = as.numeric(`Population Density, as of 1 July (persons per square km)`), 
         median_age = as.numeric(`Median Age, as of 1 July (years)`), 
         pop_growth_rate = as.numeric(`Population Growth Rate (percentage)`), 
         fertility_rate = as.numeric(`Total Fertility Rate (live births per woman)`), 
         death_rate = as.numeric(`Crude Death Rate (deaths per 1,000 population)`), 
         life_expect = as.numeric(`Life Expectancy at Birth, both sexes (years)`), 
         migration = as.numeric(`Net Migration Rate (per 1,000 population)`),
         country_un = `Region, subregion, country or area *`) %>%
  dplyr::select(-`Total Population, as of 1 July (thousands)`, -`Population Density, as of 1 July (persons per square km)`, -`Median Age, as of 1 July (years)`, -`Population Growth Rate (percentage)`, -`Total Fertility Rate (live births per woman)`, -`Crude Death Rate (deaths per 1,000 population)`, -`Life Expectancy at Birth, both sexes (years)`, -`Net Migration Rate (per 1,000 population)`,-`Region, subregion, country or area *`)

#Summary of United Nations data
summary(u_nations)
head(u_nations)

#Virtually no missing data
gg_miss_var(u_nations)

countries_2 <- u_nations %>% dplyr::select(CODE, CODE2)
typeof(countries_2)
#Saving data to the processed data folder
saveRDS(u_nations, file = here("data","processed-data","u_nations.rds"))

#Getting the total population so I can convert COVID burden from number of cases/deaths to number of cases/deaths per 100,000 persons
u_nations_total_pop <- u_nations_countries %>%
  filter(Year == 2020 | Year == 2021 | Year == 2022) %>%
  dplyr::select(CODE, CODE2, Year, `Total Population, as of 1 July (thousands)`) %>%
  mutate(total_pop = as.numeric(`Total Population, as of 1 July (thousands)`)) %>% 
  dplyr::select(-`Total Population, as of 1 July (thousands)`) %>%
  pivot_wider(
    names_from = Year, 
    values_from = total_pop,
    names_prefix = "total_pop_"
  ) %>%
  rename(
    total_pop_2020 = total_pop_2020,
    total_pop_2021 = total_pop_2021,
    total_pop_2022 = total_pop_2022
  ) %>%
  mutate(total_pop_20_22 = (total_pop_2020 + total_pop_2021 + total_pop_2022)/100)

#Summary of United Nations data
summary(u_nations_total_pop)
head(u_nations_total_pop)

```

#### COVID19 data 
This data source includes weekly COVD case count data on 239 countries. The data set includes 440640 rows, but about half have missing data on new case counts. Completeness of data will have to be reviewed once the data are restricted to countries with vaccination coverage estimates in the years of interest.
```{r}
#Importing HPV vaccination program characteristics
covid_raw <- read.csv(here("data","raw-data","WHO-COVID-19-global-daily-data.csv"), header = TRUE)

#There are missing data, year of introduction is missing for about a third of countries
gg_miss_var(covid_raw)
#Gives total number of countries in the dataset
nrow(covid_raw)
#Getting the number of countires - 239
num_countries <- table(covid_raw$Country_code)
dim(num_countries)

summary(covid_raw)
head(covid_raw)

#Have to rename the 2 character country code for merging
covid_prelim <- covid_raw %>%
  rename(CODE2 = Country_code)

#Restricting to countries of interest
covid <- countries_2 %>% left_join(covid_prelim, by = join_by(CODE2))  %>% left_join(u_nations_total_pop, by=join_by(CODE2)) %>%
  mutate(year_covid = substr(Date_reported, 1, 4)) %>% #Extracting the year from the date of data submission 
  dplyr::select(-CODE.x) %>%
  rename(CODE = CODE.y) %>%
  filter(year_covid == 2020 | year_covid == 2021 | year_covid == 2022 ) %>%
  arrange(CODE2, year_covid) %>%
  group_by(CODE2, year_covid) %>%
  slice_tail(n = 1) %>% #Taking last row in a given year per country
  filter(year_covid == 2022 ) %>% #restricted to 2022
  dplyr::select(CODE, CODE2, year_covid, Date_reported, Country, Cumulative_cases, Cumulative_deaths, total_pop_20_22) %>% #dropping unnecessary variables 
  mutate(Cumulative_case_rate = Cumulative_cases / total_pop_20_22,
         Cumulative_death_rate = Cumulative_deaths / total_pop_20_22) %>%
  dplyr::select(-Cumulative_cases, -Cumulative_deaths)

head(covid)
#Confirming all dates of submission were on 12-31-22
table(covid$Date_reported)

#Saving data to the processed data folder
saveRDS(covid, file = here("data","processed-data","covid.rds"))

# Confirmed that cumulative cases was the sum of the new_cases on all days prior
# covid_year_total <- covid %>%
#   group_by(CODE2, year_covid) %>%
#   summarise(total_cases = sum(New_cases, na.rm = TRUE), .groups = "drop")
# 
#   covid_combo <- covid %>% left_join(covid_year_total, by = join_by(CODE2, year_covid))
#   
#   summary(covid_combo)
# head(covid_combo)
# 
# last_entry <- covid_combo %>%
#   group_by(CODE2, year_covid) %>%
#   slice_tail(n = 1) %>%
#   filter(year_covid == 2020 | year_covid == 2021 | year_covid == 2022 ) %>%
#   dplyr::select(CODE2, year_covid, total_cases, Country, New_cases, Cumulative_cases)
# print(last_entry)
```

## Combining all datasets
Above, HPV vaccination coverage and vaccine program characteristic data were combined. Here, I'm combining that data with the United Nations demographic characteristics and COVID burden.

```{r}
#Joining United Nations, COVID, and HPV coverage and vaccination program data
project_data_pre <- combo %>% left_join(u_nations, by = join_by(CODE)) %>% left_join(covid, by = join_by(CODE))

#Data checks
head(project_data_pre)

#confirming the rows merged correctly across datasets by comparing the country name in each dataset
name_mismatch <- project_data_pre %>%
  filter(NAME != COUNTRYNAME | NAME != country_un | NAME != Country | COUNTRYNAME != country_un | COUNTRYNAME != Country | Country != country_un)  
print(name_mismatch)
#4 countries did not have perfectly matching names across the datasets, but all differences were due to formatting (eg., abbreviations or punctuation), confirming the merge was successful

#Dropping all extra versions of country
project_data <- project_data_pre %>%
  dplyr::select(-COUNTRYNAME, -NAME, -country_un, -CODE2.x, -Date_reported) %>%
  rename(CODE2 = CODE2.y) %>%
  mutate(coverage_diff = coverage_2022 - coverage_2019) #Creating outcome variable of difference in vaccination coverage between 2022 and 2019

#Looking at missingness for each included variable
gg_miss_var(project_data)
print(project_data %>% filter(is.na(WB_INCOME)))
#One country (Cook Islands) is missing on country income, all other data are complete

#Histogram of the outcome (difference in HPV vaccination coverage between 2022 and 2019)
hist(project_data$coverage_diff)

#Saving data to the processed data folder
saveRDS(project_data, file = here("data","processed-data","project_data.rds"))
```

### 2025-02-20 products/manuscript/Manuscript.qmd
Part 2 Commit 1 - finished data cleaning, added eda, updated READMEs, deleted unneeded files

--- 2025-01-29 products/manuscript/Manuscript.qmd
@@
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
@@
 library(openxlsx)
+library(naniar)
+library(gtsummary)
+library(flextable)
+library(gtExtras)
+library(officer)
 ```
@@
 
-* Rayleen Lewis$^{1}$ (ORCID: 0000-0000-1234-5678)
+* Rayleen Lewis$^{1}$
 
@@
 # Summary/Abstract
-_Write a summary of your project._
 
@@
 ## General Background Information
-_Provide enough background on your topic that others can understand the why and how of your analysis_
 
-## Description of data and data source
-_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
+### Description of data and data source
+The difference in HPV vaccination coverage between 2022 and 2019 will be modeled using multivariable regression techniques, including multivariable linear regression _other information on new modeling techniques will be added_.
+Cumulative number of COVID cases per 100,000 persons as of December 31, 2022 was the explanatory variable.
+HPV vaccination program characteristics, country income information, and country-level demographics will be considered for adjustment.
+HPV vaccination coverage, HPV vaccine program characteristics, and cumulative number of COVID cases were extracted from publicly available data from the World Health Organization.
+Country sociodemographic information was publicly available from the United Nations.
+All information was collected at the country level, and data were restricted to countries that had HPV vaccination coverage estimates for both 2019 and 2022.
 
-## Questions/Hypotheses to be addressed
-The objective of this work is to determine if HPV vaccination coverage changed between 2019 and 2022 among countries offering HPV vaccination as of 2019, and if so, are these changes correlated with COVID-19 burden in 2020-2021.
-I hypothesize that HPV vaccination coverage was lower in 2022 than 2019 among countries with high burden of COVID-19 in 2020-2021 and comparable in countries with a low COVID-19 burden.
+### Questions/Hypotheses to be addressed
+The objective of this work is to determine if changes in HPV vaccination coverage were correlated with COVID burden in 2020-2022.
+I hypothesize that HPV vaccination coverage was lower in 2022 than 2019 among countries with high burden of COVID-19 in 2020-2022 and comparable in countries with a low COVID-19 burden.
+
+_This is to be determined based on methods learned later in the course, but this is an initial thought on how to implement one of the methods covered in a later module._
+An additional objective of this work is to develop a classification tree analysis to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccine program and sociodemographic characteristics.
+The purpose of this classification tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV coverage if an epidemic or pandemic affects a country.
 
@@
 # Methods
-Briefly, all datasets will be reduced to include only countries with vaccination coverage estimates in both 2019 and 2022.
-Average coverage will be compared in 2022 to 2019.
-Multivariable linear regression modeling the difference in coverage between the two years as the outcome will be conducting using measures of COVID-19 burden, HPV vaccination program characteristics, country income information, and country-level demographics.
-Measure(s) of COVID-19 burden will be considered the explanatory variable of interest.
-Other characteristics will be considered for adjustment.
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
 
@@
 
-Additional country-level covariates were collected from other publicly available data sources.
-Information on country income level by year is available through the [World Bank](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups#:~:text=For%20the%20current%202025%20fiscal,those%20with%20a%20GNI%20per).
 Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)
@@
 ## Data import and cleaning
-_Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along._
-```{r, echo=FALSE}
-#world_bank <- read.xlsx("world-bank-country-income.xlsx", sheet = "Country Analytical History")
+Each data source imported and merged based on country; all datasets had variables for a country code and name. Individual datasets were restricted to the variables of interest as follows:
 
-```
+* HPV vaccination coverage
+    + HPV vaccination coverage in 2022
+    + HPV vaccination coverage in 2019
+* HPV vaccination program characteristics
+    + WHO region
+    + World Bank country income level
+    + Indicator for whether HPV vaccine is offered through a national recommendation
+    + Year of HPV vaccine program introduction
+    + Primary HPV vaccine delivery strategy (e.g., school-based)
+    + Indicator for gender neutral vaccination (i.e., female-only vaccination or females and males)
+    + Recommended number of doses (1 or 2)
+* United Nations country demographic data
+    + Total Population, as of 1 July (thousands)
+    + Population Density, as of 1 July (persons per square km)
+    + Median Age, as of 1 July (years)
+    + Population Growth Rate (percentage)
+    + Total Fertility Rate (live births per woman)
+    + Crude Death Rate (deaths per 1,000 population)
+    + Life Expectancy at Birth, both sexes (years)
+    + Net Migration Rate (per 1,000 population)
+* COVID burden through Dec. 31, 2022
+    + Cumulative number of COVID cases per 100,000 persons
+    + Cumulative number of COVID deaths per 100,000 persons
+
+The final project dataset was restricted to countries with vaccination coverage estimates in both 2019 and 2022 (N = 100).
+The outcome variable was coded as HPV vaccination coverage of at least one dose among females in 2022 minus coverage in 2019.
+Of the 100 included countries, 2 countries were missing some vaccine program characteristic information and 1 country was missing income level.
+All other countries had complete data for all variables.
 
 ## Statistical analysis
-_Explain anything related to your statistical analyses._
+#### Exploratory/Descriptive analyses
+The mean (standard deviation (sd)) and median (interquartile range (IQR)) were calculated for each continuous country characteristic (e.g., number of cumulative COVID cases).
+For categorical variables, the distribution (n, %) of each level was calculated.
+A histogram of the difference in coverage was graphed.
+Bivariate associations between each characteristic and the change in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots for continuous variables and violin plots for categorical variables.
+
+#### Multivariable modeling 
+Multivariable linear regression was performed to identify primary predictors of change in HPV vaccination coverage.
+Model selection was performed using backwards elimination; variables with the largest _p_-value were eliminated one at a time until all variables in the model had a _p_-value < 0.1.
+_Information on additionally modeling techniques will be added later_
+
+#### Classification tree analysis
+
+All data management and analysis was conducted using R (Version 2024.12.0+467). Classification tree analyses were conducted using the XXXX package.
 
@@
 
-_Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project._
+```{r, include = F, echo = F}
+table1_display=readRDS(here("results","tables","table1_tib.rds"))
+table1_tib <- as_tibble(table1_display) %>% rename(character = `Characteristic`, stat = `n/N (%), mean (sd), or median (IQR), N = 100`)
+```
 
-@tbl-summarytable shows a summary of the data.
+##### Difference in HPV vaccination coverage between 2022 and 2019
 
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
+The difference in HPV vaccination coverage between 2022 and 2019 appears normally distributed centered around 0 (i.e., no change in coverage).(@fig-Figure1) The mean (sd) difference in coverage was `r table1_tib %>%  filter(row_number() == which(character == "coverage_diff") + 1) %>% pull(stat)`; the median (IQR) was `r table1_tib %>%  filter(row_number() == which(character == "coverage_diff") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid)
+```{r}
+#| label: fig-Figure1
+#| fig-cap: "Histogram of difference in HPV vaccination coverage between 2019 and 2022"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","outcome_hist.png"))
+```
+
+{{< pagebreak >}}
 
 ```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
+#| label: tbl-Table1_coverage_covid
+#| tbl-cap: "HPV vaccination coverage and COVID burden among countries with HPV vaccination programs in 2019 and 2022"
 #| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
+knitr::include_graphics(here("results","tables","table1_cov_covid.png"))
+#knitr::knit_print(as_flex_table(table1_display))
+#table1_df=readRDS(here("results","tables","table1_df.rds"))
+#as_flextable(table1_df)
+#table1_df=readRDS(here("results","tables","table1_df.rds"))
+#knitr::knit_print(table1_df)
+#doc <- read_docx(here("products","manuscript","manuscript.docx"))
+
+# Add the gt table to the Word document
+#doc <- doc %>%
+#  body_add_gt(gt_table = table1, align = "center")
+
+# Save the Word document
+#print(doc, target = "word_table.docx")
+
+#knitr::kable(
+#  as_kable(table1_display))
+
+#table1_tbl_sum=readRDS(here("results","tables","table1_tbl_sum.rds"))
+
+#table1_tbl=readRDS(here("results","tables","try.rds"))
+#table1_tbl
+```
+
+##### COVID burden
+
+Between January 1, 2020, and December 31, 2022, the mean (sd) number of cases per 100,000 persons was `r table1_tib %>%  filter(row_number() == which(character == "Cumulative_case_rate") + 1) %>% pull(stat)`.
+The median was lower at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid) The mean 3-year COVID death rate was `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 1) %>% pull(stat)` deaths per 100,000 persons; the median was similar at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.
+
+##### Vaccination program characteristics
+The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based") + 2) %>% pull(stat)`).
+A majority of programs had a gender nuetral vaccination recommendation and a 2-dose routine recommendation (rather than a 1-dose recommendation).
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-table1_vac_program
+#| tbl-cap: "HPV vaccination program characteristics among countries with HPV vaccination programs in 2019 and 2022"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table1_vac_program.png"))
+
+```
+
+##### Sociodempgraphic characteristics
+Sociodemographic metrics were presented in @tbl-table1_demo.
+The most common WHO region was the Euro region, and the most common country income level was high-income.
+Mean life expectancy was `r table1_tib %>%  filter(row_number() == which(character == "life_expect") + 1) %>% pull(stat)`.
+
+{{< pagebreak >}}
+```{r}
+#| label: tbl-table1_demo
+#| tbl-cap: "Sociodemographic characteristics among countries with HPV vaccination programs in 2019 and 2022"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table1_demo.png"))
 ```
@@
 
-_To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
 ## Full analysis
-
-_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here._
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
 
@@
 ## Summary and Interpretation
-_Summarize what you did, what you found and what it means._
 
 ## Strengths and Limitations
-_Discuss what you perceive as strengths and limitations of your analysis._
 
 ## Conclusions
-_What are the main take-home messages?_
-
-_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
 
@@
 # References
-
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

### 2025-03-19 Code chunks for reference.qmd
Part 3 Commit 1

@@
 Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)
+
+# Package management
+
+It is recommended to use [renv](https://rstudio.github.io/renv/index.html) to manage R packages and increase chances of future reproducibility.
+This is required if you are using the template as part of a research project for our group.
+Otherwise, you can decide to implement `renv` or not.
+This can happen at any stage, though earlier in the project is generally better.
+
+If you plan to use `renv`, start by reading [the introduction to `renv` article](https://rstudio.github.io/renv/articles/renv.html) so you know how to use it.

### 2025-03-19 code/eda-code/eda.qmd
Part 3 Commit 1

@@
 library(gtsummary)
+library(procs)
 ```
@@
 saveRDS(means_table, file = meanstable_file)
+
+### Checks to determine if there are any countries with extremely low coverage estimates that seem implausible
+proc_means(project_data, var = c("coverage_2019","coverage_2022"), stats = c("N","NMISS","MEAN","STD","MEDIAN","Q1","Q3","MIN","MAX"), options = v(maxdec = 2))
+
+project_data %>% filter(coverage_2019 < 20 | coverage_2022 < 20) %>% dplyr::select(Country, coverage_diff, coverage_2019, coverage_2022)
+#Several countries have coverage estimates that seem to be data quality issues (e.g., Botswana, Gambia, Cook Islands), but there's no way to actually know if coverage is actually that low or if coverage estimates are extremely off - I'll leave data as is for now.
 

### 2025-03-19 products/manuscript/Manuscript.qmd
Part 3 Commit 1

@@
 ## Statistical analysis
-#### Exploratory/Descriptive analyses
+### Exploratory/Descriptive analyses
 The mean (standard deviation (sd)) and median (interquartile range (IQR)) were calculated for each continuous country characteristic (e.g., number of cumulative COVID cases).
@@
 
-#### Multivariable modeling 
-Multivariable linear regression was performed to identify primary predictors of change in HPV vaccination coverage.
-Model selection was performed using backwards elimination; variables with the largest _p_-value were eliminated one at a time until all variables in the model had a _p_-value < 0.1.
-_Information on additionally modeling techniques will be added later_
+### Multivariable modeling 
+Multivariable linear regression was performed to assess the association between difference in HPV vaccination coverage and cumulative number of COVID cases, a measure of COVID burden.
+An unadjusted model and fully adjusted model (adjusted for all country level demographic and vaccination program characteristics) were performed.
+Additionally, models adjusted for each country level characteristic were estimated.
+For each model, the effect of COVID case count on difference in HPV vaccination coverage between 2019-2022 (beta , standard error, p-value from t-test with a null hypothesis of beta = 0) and model performance metrics (R-squared, AIC, BIC) were extracted and compared.
+Selection of a final model was based on subject matter expertise, which model had the highest R-squared, and which model had the lowest AIC and/or BIC.
+Model parameters were presented for the final selected model.
 
-#### Classification tree analysis
+### Classification tree analysis
 
@@
 
-##### Difference in HPV vaccination coverage between 2022 and 2019
+### Difference in HPV vaccination coverage between 2022 and 2019
 
@@
 
-##### COVID burden
+### COVID burden
 
@@
 
-##### Vaccination program characteristics
-The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based") + 2) %>% pull(stat)`).
+### Vaccination program characteristics
+The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based")) %>% pull(stat)`).
 A majority of programs had a gender nuetral vaccination recommendation and a 2-dose routine recommendation (rather than a 1-dose recommendation).
@@
 
-##### Sociodempgraphic characteristics
+### Sociodempgraphic characteristics
 Sociodemographic metrics were presented in @tbl-table1_demo.
@@
 ## Basic statistical analysis
+### Bivariate associations between the difference in HPV vaccination coverage between 2019 and 2022 and COVID burden 
+There appears to be linear association between cumulative COVID cases and the difference in HPV vaccination coverage between 2019 and 2022 (@fig-Figure2).
+Of note, there is a lot of variation in the coverage difference among countries with lower numbers of cumulative COVID cases.
+
+```{r}
+#| label: fig-Figure2
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID cases per 100,000"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","scatter_case_covdiff.png"))
+
+```
+
+A similar association was seen when assessing cumulative COVID deaths (@fig-Figure3).
+Because these variables are both measures of COVID burden and they exhibit similar patterns, the remainder of the analysis is focused on COVID case burden.
+
+```{r}
+#| label: fig-Figure3
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID deaths per 100,000"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","scatter_death_covdiff.png"))
+```
+
+### Bivariate associations between the each country demographic and vaccination program characteristic and the difference in HPV vaccination coverage between 2019 and 2022  
+The distribution of the difference in vaccination coverage by each country level demographic variable can be seen in @fig-Figure4.
+For each WHO region, the distribution of change in coverage centers around 0, with the SEAR and EUR showing a more narrow distribution.
+Similarly, the distribution of coverage difference was centered around 0 regardless of country income level, and high-income countries had the tightest distribution.
+The patterns in the continuous variables are less straightforward, particularly as there appear to be potential outliers.
+Based on the figure below, total population may have a linear, positive association with coverage difference, while life expectancy and median age show a negative quadratic association and fertility rate may show a negative quadratic association.
+
+```{r}
+#| label: fig-Figure4
+#| fig-cap: "Distribution of difference in HPV vaccination coverage between 2019 and 2022 by each country demographic variable"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","demo_fig.png"))
+
+```
+
+There were some differences in the distribution of coverage difference by vaccine program characteristics (@fig-Figure5).
+Countries with a facility-based delivery strategy, gender neutral vaccination program, and 2-dose recommendation had distributions of coverage difference with a smaller portion of countries in the distribution region indicating a decrease in coverage (i.e., most of the data was centered around zero or was greater than 0).
+There was not a strong association with difference in coverage and year of association.
+
+```{r}
+#| label: fig-Figure5
+#| fig-cap: "Associations between vaccine program characteristics and difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID deaths per 100,000"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","bivar_vac.png"))
+```
+
+## Preliminary modeling
+As a preliminary modeling step, beta estimates, standard errors, and p-values for the effect of cumulative COVID cases on the difference in coverage are presented in @tbl-modelcomp.
+Each performance metric has been shaded using a sliding color scale; a light color indicates worse performance, darker color indicates better performance, and the "best" performing model is indicated using bold text.
+The fully adjusted model (adjusted for all country level demographic and vaccination program characteristics) had the highest R^2^ and lowest AIC.
+The model adjusted for country income level had the lowest BIC.
+Therefore, preliminary modeling will focus on the fully adjusted model.
+However, none of these models are a particularly great fitting model, and given that the beta is switching directions (positive/negative) and very close to zero, there does not appear to be a strong association between COVID burden and difference in coverage regardless of adjustment.
+```{r}
+#| label: tbl-modelcomp
+#| tbl-cap: "Preliminary model comparison for models assessing the effect of cumulative COVID cases on the difference in HPV vaccination coverage"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table_model_adj_format.png"))
+
+```
+
+Estimates from the fully adjusted model are shown in @tbl-fully_adj.
+There was not a significant association between cumulative number of COVID cases and the difference in HPV vaccination coverage between 2019-2022.
+After adjusting for all other characteristics, the difference in coverage was significantly higher and positive (indicating increase in coverage between 2019 and 2022) for all WHO regions, except the Eastern Mediterranean Region, compared to the African Region.
+```{r}
+#| label: tbl-fully_adj
+#| tbl-cap: "Model parameters for fully adjusted model assessing the effect of cumulative COVID cases on the difference in HPV vaccination coverage"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","lmtablefull_gt.png"))
+
+```
 

### 2025-03-19 code/analysis-code/statistical-analysis.R

@@ -8,49 +8,428 @@
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
-
+library(tidyverse)
+library(patchwork)
+library(dplyr)
+library(tidyr)
+library(gt)
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
+data_location <- here::here("data","processed-data","project_data.rds")
 
 #load data. 
-mydata <- readRDS(data_location)
+project_data <- readRDS(data_location)
 
 
 ######################################
 #Data fitting/statistical analysis
 ######################################
 
+
+
+
+
+
+#Violin plots of the outcome by each categorical variable
+demo1 <- project_data %>%
+  ggplot(aes(coverage_diff, WHO_REGION)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "WHO region")
+
+demo2 <- project_data %>%
+  ggplot(aes(coverage_diff, WB_INCOME)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Country income level")
+
+demo3 <- project_data %>%
+  ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Total population (in thousands)")
+
+demo4 <- project_data %>%
+  ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Population density (persons per square km)")
+
+#Building 4 panel figure of country demographics
+bivar_demo_1 <- (demo1 | demo2) / (demo3 | demo4)
+bivar_demo_1_loc <- here("results","figures","bivar_demo_1.png")
+# Save the plot created as a PNG
+ggsave(bivar_demo_1_loc, plot = bivar_demo_1, width = 8, height = 6, dpi = 300)
+
+
+demo5 <- project_data %>%
+  ggplot(aes(coverage_diff, median_age)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Median age of population")
+
+demo6 <- project_data %>%
[…]

### 2025-04-04 code/eda-code/eda_for_manuscript.qmd
Part 4 Commit 1

@@
 ```{r}
-#histogram of differnce in coverage
+#histogram of difference in coverage
 outcome_hist <- ggplot(project_data, aes(x = coverage_diff)) +

### 2025-04-04 products/manuscript/Manuscript.qmd
Part 4 Commit 1

--- 2025-02-20 products/manuscript/Manuscript.qmd
@@
 
-* Rayleen Lewis$^{1}$
+-   Rayleen Lewis$^{1}$
 
@@
 
-$\land$ Corresponding author: rml20258@uga.edu
+$\land$ Corresponding author: rml20258\@uga.edu
 
-$\dagger$ Disclaimer: The opinions expressed in this article are the
-author's own and don't reflect their employer or University.
+$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer or University.
 
@@
 
-# Summary/Abstract
+# Abstract
+
+## Background:
+
+Human papillomavirus vaccines have been available since 2006, and as of 2019, 108 countries had introduced HPV vaccine into their national immunization programs.
+In 2020, immunization services globally experienced disruptions due to the COVID-19 pandemic.
+My objective was to assess the association between country-level COVID burden and changes in HPV vaccination coverage between 2019 and 2022.
+
+## Methods:
+
+Data on annual coverage, cumulative number of COVID cases between Jan.
+1, 2020, and Dec.
+31, 2022, country income level, vaccination program characteristics, and sociodemographic information were publicly available from the World Health Organization and United Nations.
+Multivariable linear regression models, modeling the difference in 2022 and 2019 HPV vaccination coverage, were performed with number of cumulative COVID cases as the primary predictor; additional country vaccination program and sociodemographic characteristics were considered for adjustment.
+Analyses were restricted to complete cases (N = 98).
+LASSO regression and regression tree models were also considered.
+Models were compared using R^2^ values.
+
+## Results:
+
+There was a weak, positive, linear association between the difference in HPV vaccination coverage and number of cumulative COVID cases (𝛽(se) = 0.000769 (0.00036), p-value = 0.034).
+Variability in the coverage difference was highest for countries with the lowest COVID burden and decreased as COVID burden increased.
+After adjustment for all vaccination program and sociodemographic characteristics, coverage difference and COVID burden were not associated (beta = -0.000065 (0.00064), p-value = 0.919).
+COVID burden was an important predictor of difference in coverage among countries with a median resident age of \<26 years.
+
+## Conclusions:
+
+After accounting for additional country-level characteristics, COVID burden was not associated with the change in coverage between 2019 and 2022 for countries overall.
+There is high variability between countries in the ability and methods used to assess both HPV vaccination coverage and COVID burden.
+It is unclear if the null association observed would be replicated if countries used similar methodology for assessing national health metrics.
 
@@
 
-## General Background Information
+Human papillomavirus (HPV) is the most common sexually transmitted infection globally.[@sanjoseWorldwidePrevalenceGenotype2007] There are about 40 mucosal HPV types, and at least 14 are considered high-risk, or oncogenic, and may lead to cancer.[@lewisEstimatedPrevalenceIncidence2021] Globally, cervical cancer is the most common HPV-attributable cancer and is the 4th most common cancer among women.
+In 2022, almost 350,000 women died of cervical cancer, with over 90% of these deaths occurring among women who live in low and middle income countries.[@worldhealthorganizationGlobalStrategyAccelerate2020]
 
-### Description of data and data source
-The difference in HPV vaccination coverage between 2022 and 2019 will be modeled using multivariable regression techniques, including multivariable linear regression _other information on new modeling techniques will be added_.
-Cumulative number of COVID cases per 100,000 persons as of December 31, 2022 was the explanatory variable.
-HPV vaccination program characteristics, country income information, and country-level demographics will be considered for adjustment.
-HPV vaccination coverage, HPV vaccine program characteristics, and cumulative number of COVID cases were extracted from publicly available data from the World Health Organization.
-Country sociodemographic information was publicly available from the United Nations.
-All information was collected at the country level, and data were restricted to countries that had HPV vaccination coverage estimates for both 2019 and 2022.
+Several HPV vaccines are available to prevent infections that cause most HPV-attributable cancers.
+The first vaccine product, a quadrivalent vaccine that protects against 4 HPV types, entered the market in 2006; the United States was the first country to make a national HPV vaccination recommendation.[@markowitzTenYearsHuman2018; @bruniHPVVaccinationIntroduction2021] Since then, five additional vaccine products have entered the market and over 100 countries have introduced HPV vaccination into their national immunization program.[@whoHumanPapillomavirusVaccines2022; @bruniHPVVaccinationIntroduction2021]
 
-### Questions/Hypotheses to be addressed
-The objective of this work is to determine if changes in HPV vaccination coverage were correlated with COVID burden in 2020-2022.
-I hypothesize that HPV vaccination coverage was lower in 2022 than 2019 among countries with high burden of COVID-19 in 2020-2022 and comparable in countries with a low COVID-19 burden.
-
-_This is to be determined based on methods learned later in the course, but this is an initial thought on how to implement one of the methods covered in a later module._
-An additional objective of this work is to develop a classification tree analysis to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccine program and sociodemographic characteristics.
-The purpose of this classification tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV coverage if an epidemic or pandemic affects a country.
-
-{{< pagebreak >}}
+In December 2019, cases of a severe respiratory disease increased rapidly in China.
+We now know this was the start of the COVID-19 pandemic.[@cucinotta2020] Globally, this pandemic disrupted many medical services, including immunization programs.[@casey2024] Several studies have demonstrated the country-specific impact of the COVID-19 pandemic on HPV vaccination and HPV-related disease.[@ivanus2021; @gountas2023; @silva2022; @rao2022; @daniels2021] Additionally, a global assessment of the effect of the pandemic on HPV vaccination coverage has been performed.[@casey2024] However, these analyses focused on pre- and post-pandemic comparisons rather than considering the magnitude of COVID burden in the country.
+To address this gap, my objective was to assess the association between country-level COVID burden and changes in HPV vaccination coverage between 2019 and 2022.
+I hypothesized that decreases in HPV vaccination coverage between 2022 and 2019 were greater among countries with high burden of COVID-19 in 2020-2022.
+A secondary objective of this work was to develop a regression tree to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccine program and sociodemographic characteristics.
+The purpose of this regression tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV coverage if a new epidemic or pandemic were to affect a country.
 
@@
 
-## Data aquisition
-Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).
-Briefly, coverage data reflect administrative and official HPV vaccination coverage reported annually through the WHO/UNICEF Joint Reporting Form on Immunization.
-HPV Dashboard include country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccine schedule, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (target males and females), WHO region, and year-specific coverage from 2010 to 2023.
+## Data acquisition
 
-Country-specific weekly COVID-19 case and death counts are available from starting in 1/4/2020 through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o).
-In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID-19 data.
+Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).[@HPVdashboard] Briefly, coverage data reflect administrative and official HPV vaccination coverage reported annually through the WHO/UNICEF Joint Reporting Form on Immunization.
+The HPV Dashboard includes country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccination recommendation, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (i.e., targets males and females), WHO region, and year-specific coverage from 2010 to 2023.
 
-Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format)
+Country-specific weekly COVID-19 case and death counts are available starting on January 4, 2020, through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o).[@COVID19Dashboard] In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID-19 data.
+
+Country-level demographics including life expectancy of females at birth and population size were available from the [United Nations](https://population.un.org/wpp/downloads?folder=Standard%20Projections&group=CSV%20format).[@UNWorldPopulationProspects]
 
 All data sources are publicly available for download and were downloaded on January 26-27, 2025.
+Each data source was imported and merged based on a country code identifier, which was present in all datasets along with country name.
 
-## Data import and cleaning
-Each data source imported and merged based on country; all datasets had variables for a country code and name. Individual datasets were restricted to the variables of interest as follows:
+The data were first restricted to countries with vaccination coverage estimates in both 2019 and 2022.
+The outcome variable was coded as HPV vaccination coverage of at least one dose among females in 2022 minus coverage in 2019.
+Of the 100 countries with coverage information from both years, 2 countries were missing some vaccine program characteristic information, one of which was also missing income level.
+All other countries had complete data for all variables.
+Complete cases were used for all analyses (N = 98).
 
-* HPV vaccination coverage
-    + HPV vaccination coverage in 2022
-    + HPV vaccination coverage in 2019
-* HPV vaccination program characteristics
-    + WHO region
-    + World Bank country income level
-    + Indicator for whether HPV vaccine is offered through a national recommendation
-    + Year of HPV vaccine program introduction
-    + Primary HPV vaccine delivery strategy (e.g., school-based)
-    + Indicator for gender neutral vaccination (i.e., female-only vaccination or females and males)
-    + Recommended number of doses (1 or 2)
-* United Nations country demographic data
-    + Total Population, as of 1 July (thousands)
-    + Population Density, as of 1 July (persons per square km)
-    + Median Age, as of 1 July (years)
-    + Population Growth Rate (percentage)
-    + Total Fertility Rate (live births per woman)
-    + Crude Death Rate (deaths per 1,000 population)
-    + Life Expectancy at Birth, both sexes (years)
-    + Net Migration Rate (per 1,000 population)
-* COVID burden through Dec. 31, 2022
-    + Cumulative number of COVID cases per 100,000 persons
-    + Cumulative number of COVID deaths per 100,000 persons
+## Country-level characteristics
 
-The final project dataset was restricted to countries with vaccination coverage estimates in both 2019 and 2022 (N = 100).
-The outcome variable was coded as HPV vaccination coverage of at least one dose among females in 2022 minus coverage in 2019.
-Of the 100 included countries, 2 countries were missing some vaccine program characteristic information and 1 country was missing income level.
-All other countries had complete data for all variables.
+COVID burden between January 4, 2020, and December 31, 2022, was defined as the cumulative number of COVID cases per 100,000 persons.
+Additionally, cumulative number of COVID deaths per 100,000 was also calculated.
+
+Vaccination program characteristics include WHO region, country income level, presence of a national vaccination recommendation, year of HPV vaccination program introduction, primary vaccine delivery strategy (e.g., school-based), whether the program offers gender neutral vaccination (i.e., female-only vaccination or females and males), and recommended number of doses.
+
+Country-level demographic characteristics include total population (thousands), population density, median age of residents, population growth rate, total fertility rate, crude death rate, life expectancy at birth, and net migration rate.
+
+Additional details for each characteristic can be found in the Supplementary Material.
 
 ## Statistical analysis
-#### Exploratory/Descriptive analyses
+
+All data management and statistical analyses were performed using R version 4.4.2 (2024-10-31) within RStudio Version 2024.12.0+467.
+
+### Descriptive analyses
+
+Univariate analyses of the outcome and each country-level characteristic were performed.
+A histogram of the difference in coverage was assessed to determine the approximate distribution of the outcome.
 The mean (standard deviation (sd)) and median (interquartile range (IQR)) were calculated for each continuous country characteristic (e.g., number of cumulative COVID cases).
 For categorical variables, the distribution (n, %) of each level was calculated.
-A histogram of the difference in coverage was graphed.
-Bivariate associations between each characteristic and the change in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots for continuous variables and violin plots for categorical variables.
+Bivariate associations between each characteristic and the change in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots (with a LOESS line of best fit for visualization purposes only) for continuous variables and violin plots for categorical variables.
 
-#### Multivariable modeling 
-Multivariable linear regression was performed to identify primary predictors of change in HPV vaccination coverage.
-Model selection was performed using backwards elimination; variables with the largest _p_-value were eliminated one at a time until all variables in the model had a _p_-value < 0.1.
-_Information on additionally modeling techniques will be added later_
+### Multivariable modeling
 
-#### Classification tree analysis
+Multivariable linear regression was performed using the lm function to assess the association between the difference in HPV vaccination coverage and cumulative number of COVID cases, a measure of COVID burden.
+An unadjusted model and fully adjusted model (adjusted for all country level demographic and vaccination program characteristics) were performed.
+Additionally, models adjusted for each country level characteristic individually were performed.
+For each model, the effect of COVID case count on the difference in HPV vaccination coverage between 2019-2022 (beta, standard error, p-value from t-test with a null hypothesis of beta = 0) and model performance metrics (R-squared, AIC, BIC) were extracted and compared.
+Selection of a final model was based on subject matter expertise, which model had the highest R-squared, and which model had the lowest AIC and/or BIC.
+Model parameters were presented for the final selected model.
+Because there are many potential predictors, LASSO regression was also performed to reduce the risk of overfitting.
+LASSO regression was fit for a grid of penalty values from 1E-5 to 1E2 on a log scale using the glmnet engine and tidymodels framework.
+The highest tuning value without a reduction in R^2^ was selected as the final tuning value.
+Beta estimates and R^2^ values from the final multivariable linear regression and LASSO regression models were compared.
+Diagnostic plots (e.g., predicted vs observed) for both models were visually inspected to assess model fit.
 
-All data management and analysis was conducted using R (Version 2024.12.0+467). Classification tree analyses were conducted using the XXXX package.
+### Regression tree analysis
+
+Regression tree analyses were conducted using the rpart package.
+This package builds a regression tree using a two step process.
+First, the variable that best splits the data into two groups is identified.
+The process is repeated on the two resulting groups.
+This is repeated until subgroups either reach the minimum sample size or the model cannot be improved by additional variable splits.
+Second, cross validation is used to trim leaves from the tree to reduce the risk of overfitting.
+All characteristics were included as potential predictors in the regression tree analysis.
+R^2^ was estimated for the best fitting model for comparison with the multivariable linear and LASSO regressions.
 
@@
 
-## Exploratory/Descriptive analysis
+## Descriptive analyses
 
@@
 
-##### Difference in HPV vaccination coverage between 2022 and 2019
+### Univariate analyses
+
+#### Difference in HPV vaccination coverage between 2022 and 2019
 
 The difference in HPV vaccination coverage between 2022 and 2019 appears normally distributed centered around 0 (i.e., no change in coverage).(@fig-Figure1) The mean (sd) difference in coverage was `r table1_tib %>%  filter(row_number() == which(character == "coverage_diff") + 1) %>% pull(stat)`; the median (IQR) was `r table1_tib %>%  filter(row_number() == which(character == "coverage_diff") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid)
+
 ```{r}
 #| label: fig-Figure1
-#| fig-cap: "Histogram of difference in HPV vaccination coverage between 2019 and 2022"
+#| fig-cap: "Histogram of difference in HPV vaccination coverage between 2022 and 2019"
 #| echo: FALSE
@@
 
-##### COVID burden
+#### COVID burden
 
 Between January 1, 2020, and December 31, 2022, the mean (sd) number of cases per 100,000 persons was `r table1_tib %>%  filter(row_number() == which(character == "Cumulative_case_rate") + 1) %>% pull(stat)`.
-The median was lower at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid) The mean 3-year COVID death rate was `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 1) %>% pull(stat)` deaths per 100,000 persons; the median was similar at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.
+The median (IQR) was lower at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid) The mean 3-year COVID death rate was `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 1) %>% pull(stat)` deaths per 100,000 persons; the median was similar at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 2) %>% pull(stat)`.
 
-##### Vaccination program characteristics
-The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based") + 2) %>% pull(stat)`).
+#### Vaccination program characteristics
+
+The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based")) %>% pull(stat)`).
 A majority of programs had a gender nuetral vaccination recommendation and a 2-dose routine recommendation (rather than a 1-dose recommendation).
@@
 
-##### Sociodempgraphic characteristics
+#### Sociodemographic characteristics
+
 Sociodemographic metrics were presented in @tbl-table1_demo.
 The most common WHO region was the Euro region, and the most common country income level was high-income.
-Mean life expectancy was `r table1_tib %>%  filter(row_number() == which(character == "life_expect") + 1) %>% pull(stat)`.
+Mean life expectancy was `r table1_tib %>%  filter(row_number() == which(character == "life_expect") + 1) %>% pull(stat)` years.
 
 {{< pagebreak >}}
+
 ```{r}
@@
 
-## Basic statistical analysis
+### Bivariate analyses
 
-## Full analysis
+#### Bivariate associations between the difference in HPV vaccination coverage between 2022 and 2019 and COVID burden
+
+There was a weak linear association between cumulative COVID cases and the difference in HPV vaccination coverage between 2019 and 2022 (@fig-Figure2).
+Of note, there is a lot of variation in the coverage difference among countries with fewer cumulative COVID cases per 100,000 persons.
+
+```{r}
+#| label: fig-Figure2
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID cases per 100,000"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","scatter_case_covdiff.png"))
+
+```
+
+A similar association was seen when assessing cumulative COVID deaths (Supplementary material).
+Because these variables are both measures of COVID burden and they exhibited similar patterns, the remainder of the analysis focused on COVID case burden.
+
+#### Bivariate associations between the each country demographic and vaccination program characteristic and the difference in HPV vaccination coverage between 2022 and 2019
+
+The distribution of the difference in vaccination coverage by each country-level demographic variable can be seen in @fig-Figure4.
+For each WHO region, the distribution of difference in coverage centers around 0, with the SEAR and EUR showing a more narrow distribution.
+Similarly, the distribution of coverage difference was centered around 0 regardless of country income level, and high-income countries had the tightest distribution.
+The patterns in the continuous variables are less straightforward, particularly as there appear to be potential outliers.
+Based on the figure below, total population may have a linear, positive association with coverage difference, while life expectancy and median age show a negative quadratic association and fertility rate may show a positive quadratic association.
+
+```{r}
+#| label: fig-Figure4
+#| fig-cap: "Distribution of difference in HPV vaccination coverage between 2022 and 2019 by each country demographic variable"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","demo_fig.png"))
+
+```
+
+There were some differences in the distribution of coverage difference by vaccine program characteristics (@fig-Figure5).
+Countries with a facility-based delivery strategy, gender neutral vaccination program, and 2-dose recommendation had distributions of difference in coverage with a smaller portion of countries in the distribution region indicating a decrease in coverage (i.e., most of the data was centered around zero or was greater than 0).
+There was not a strong association with difference in coverage and year of implementation.
+
+```{r}
+#| label: fig-Figure5
+#| fig-cap: "Associations between vaccine program characteristics and difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID deaths per 100,000"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","bivar_vac.png"))
+```
+
+## Multivariable modeling
+
+### Model building and selection
+
+#### Multivariable linear regression
+
+```{r, include = F, echo = F}
+#Loading the betas from the linear and LASSO regression models so I can call the output directly in text
+prelim_lm_betas <- tibble(readRDS(here("results", "tables", "table_model_adj_format.rds")))
+prelim_lm_betas <- tibble(prelim_lm_betas[[1]][[1]])
+#x <- prelim_lm_betas[[1]][[1]]
+```
+
+For the unadjusted and all adjusted multivariable linear regression models, COVID burden beta estimates, standard errors, p-values for the effect of cumulative COVID cases on the difference in coverage, and performance metrics (R^2^, AIC, BIC) are presented in @tbl-modelcomp.
+Each performance metric has been shaded using a sliding color scale; a light color indicates worse performance, darker color indicates better performance, and the "best" performing model is indicated using bold text.
+The fully adjusted model (adjusted for all country-level demographic and vaccination program characteristics) had the highest R^2^.
+The model adjusted for median resident age had the lowest AIC and BIC.
+The association between change in coverage and COVID burden were similar in the fully adjusted model (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Full model")) %>% pull(beta) %>% format(scientific = FALSE)`) and model only adjusted for median age (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Median age")) %>% pull(beta) %>% format(scientific = FALSE)`) (@tbl-modelcomp, see supplementary material for all model parameters for median age adjusted model).
+Preliminary modeling will focus on the fully adjusted model.
+However, none of these models fit particularly well based on the low R^2^ values and diagnostic plots in @fig-lm-diagnostics.
+Based on the predicted vs observed plot, many observations with relatively high or low observed coverage difference values are further from the line of perfect fit.
+Additionally, there appears to be a slight pattern in the residuals; relatively small predicted values are more likely to have negative residuals and moderate values are more likely to have large positive values.
+Given that the beta is switching directions (positive/negative) based on the adjustment variables and is very close to zero (@tbl-modelcomp), there does not appear to be a strong association between COVID burden and difference in coverage regardless of adjustment.
+
+```{r}
+#| label: tbl-modelcomp
+#| tbl-cap: "Preliminary model comparison for models assessing the effect of cumulative COVID cases on the difference in HPV vaccination coverage"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","table_model_adj_format.png"))
+
+```
+
+```{r}
+#| label: fig-lm-diagnostics
+#| fig-cap: "Diagnostic plots for the fully adjusted multivariable linear regression model: predicted vs observed values (left panel) and predicted values vs residuals (right panel)"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","full_lm_diagnostics.png"))
+
+```
+
+#### LASSO regression
+
+Because of the large number of coefficients in the full model, LASSO regression model was also performed for regularization.
+A grid of penalty values from 1E-5 to 1E2 on a log scale was considered for tuning.
+@Fig-Lasso_fit shows the R^2^ value for each model tuning parameter in the left panel, and the residuals from the run with each tuning parameter in the right panel.
+Each color represents a particular run.
+R^2^ is relatively stable for tuning parameters values from 1E-5 to 1E-2; therefore, a tuning value of 1E-2 was selected for the final model.
+
+```{r}
+#| label: fig-Lasso_fit
+#| fig-cap: "LASSO regression fit diagnostic plots showing the effect of tuning parameter on model R^2^ (left panel) and residuals by predicted values for each tuning parameter (right panel)"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","lasso_fit_plot.png"))
+
+```
+
+#### Regression tree analysis
+
+Median age was the variable identified for the first node, splitting the countries based on median age \< 26 years.
+For countries with a median age \< 26, countries were further split by cumulative COVID cases per 100,000 persons (\>=155, \<155).
+Of these, countries with a higher COVID burden (n = 17) had a 27 percentage point decrease in HPV vaccination coverage on average; countries with a lower COVID burden (n = 7) had an average increase in coverage of 1.9 percentage points.
+For countries with a median age \>=26 years, further dichotomizations were made based on an additional median age dichotomization, population density, fertility rate, migration rate, and number of HPV doses recommended.
+
+```{r}
+#| label: fig-tree
+#| fig-cap: "Regression tree of coverage difference between 2022 and 2019 using all predictors"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","regression_tree.png"))
+
+```
+
+### Model results and comparison
+
+```{r, include = F, echo = F}
+#Loading the betas from the linear and LASSO regression models so I can call the output directly in text
+reg_betas <- readRDS(here("results", "tables", "lmtablefull_gt_ref.rds"))
+#Loading the R-sqr values from the models so I can reference them in the text
+r_sq_table <- readRDS(here("results", "tables", "r_sq_table.rds"))
+#Renaming r-squared column so this can be called in the text below, b/c the "-" is causing issues otherwise
+r_sq_table_ref <- r_sq_table %>% rename(rsq_est = `R-squared`)
+
+```
+
+Estimates from the fully adjusted model are shown in @tbl-fully_adj.
+There was not a significant association between cumulative number of COVID cases and the difference in HPV vaccination coverage between 2019-2022 using the linear model (𝛽(se) = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(estimate) %>% format(scientific = FALSE)` (`r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(std.error) %>% format(scientific = FALSE)`), *p*-value = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(p.value)`).
+After adjusting for all other characteristics, the difference in coverage was significantly higher and positive (indicating increase in coverage between 2019 and 2022) for all WHO regions, except the Eastern Mediterranean Region, compared to the African Region.
+Conclusions from the LASSO regression model were similar.
+However, some beta estimates other than COVID burden changed magnitude and/or direction (e.g., population growth rate, low income) compared to the linear regression model.
+
+```{r}
+#| label: tbl-fully_adj
+#| tbl-cap: "Model parameters for fully adjusted model assessing the effect of cumulative COVID cases on the difference in HPV vaccination coverage"
+#| echo: FALSE
+knitr::include_graphics(here("results","tables","lmtablefull_gt.png"))
+
+```
+
+Based on the regression tree (@fig-tree), median age of residents was the most important characteristic for grouping countries based on their coverage differences between 2022 and 2019.
+COVID burden was an important predictor among countries with a median age of \< 26 years for persons residing in the country.
+
+Of the four models considered, the regression tree had the largest R^2^ value (`r r_sq_table_ref %>% filter(row_number() == which(Model == "Regression tree")) %>% pull(rsq_est) %>% round(3)`).
+This model is also potentially the most intuitive, and the utility of estimating COVID burden is apparent in this model.
+
+```{r}
+#| label: tbl-rsq-models
+#| tbl-cap: "R^2^ comparison of models"
+#| echo: FALSE
+#r_sq_table = readRDS(here("results", "tables", "r_sq_table.rds")) - this RDS file is loaded above
+knitr::kable(r_sq_table)
+```
 
@@
 
-## Summary and Interpretation
+Overall, there was not a significant association between COVID burden and changes in HPV vaccination coverage after adjusting for other country level characteristics.
+Previous COVID research found countries with high levels of internal income inequality were disproportionately impacted by COVID.[@suIncomeInequalityDisease2022] The observed lack of an association in the present analysis may be due to masked regional or local differences since only national level data are available.
 
-## Strengths and Limitations
+While all countries use the same case reporting form to submit HPV vaccination coverage estimates to WHO, the methods used to generate the initial estimates vary widely.
+WHO does an independent assessment of each country's data to better understand if changes in estimates from year to year may be due to methodological issues in data collection or if estimates have truly changed.[@whoWHOUNICEFImmunization2024] This may be particularly problematic if reporting and COVID burden were influenced by an additional factor, such as country income level.
+To mitigate potential confounding effects, country-level demographic characteristics (e.g., median age of residents), income level, and vaccination program characteristics were considered for adjustment.
+However, residual confounding may be possible.
 
-## Conclusions
+A strength of this work is the development of the regression tree analysis that may be used by countries if faced with a pandemic in the future.
+Among countries with a younger median age of its residents, countries who are experiencing a higher burden of disease due to a pandemic may choose to prioritize  resources to minimize changes in vaccination coverage.
+Based on the results in the present study, this may be less of a priority for countries with lower pandemic disease burden or countries with older residents on average.
+
+This analysis is subject to several limitations.
+First, as discussed above, countries vary in the methods used to estimate national health indicators, including vaccination coverage.
+Second, the absolute number of countries was relatively small.
+However, these countries represent more than half of the countries reporting to the World Health Organization.
+Third, the analysis was restricted to countries who had HPV vaccination coverage estimates in 2019 and 2022, meaning only countries that had introduced HPV vaccination by 2019 were included.
+In the event of a future pandemic, it is unclear if these findings can be generalized to countries that introduced HPV vaccination after 2019 or choose not to introduce a national HPV vaccination program.
+
+## Conclusion
+
+During the first two years of the COVID-19 pandemic, many countries experienced a decrease in HPV vaccination coverage.
+Overall, COVID burden was not a strong predictor of changes in HPV vaccination coverage.
+However, for countries with a younger median age of residents, higher COVID burden was associated with larger decreases in HPV vaccination coverage, highlighting the importance of developing country-specific pandemic preparedness efforts.
 

### 2025-04-04 products/manuscript/supplement/Supplementary-Material.qmd
Part 4 Commit 1

---
title: "Supplement to Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Supplemental Information

## Table of contents

-   Code and file information
-   Country characteristic details

# Code and file information

Explain here what each code/file is and does, and in which order (if any) users need to run thing to reproduce everything.
Essentially, give a full set of instructions to re-generate everything.

# Country characteristic details
The following list provides additional details for each country-level characteristics, including data source - World Health Organization (WHO) or United Nations (UN)

-   HPV vaccination coverage (WHO)
    -   HPV vaccination coverage in 2022
    -   HPV vaccination coverage in 2019
-   HPV vaccination program characteristics (WHO)
    -   WHO region
    -   World Bank country income level
    -   Indicator for whether HPV vaccine is offered through a national recommendation
    -   Year of HPV vaccine program introduction
    -   Primary HPV vaccine delivery strategy (e.g., school-based)
    -   Indicator for gender neutral vaccination (i.e., female-only vaccination or females and males)
    -   Recommended number of doses (1 or 2)
-   United Nations country demographic data (UN)
    -   Total Population, as of 1 July (thousands)
    -   Population Density, as of 1 July (persons per square km)
    -   Median Age, as of 1 July (years)
    -   Population Growth Rate (percentage)
    -   Total Fertility Rate (live births per woman)
    -   Crude Death Rate (deaths per 1,000 population)
    -   Life Expectancy at Birth, both sexes (years)
    -   Net Migration Rate (per 1,000 population)
-   COVID burden through Dec. 31, 2022 (WHO)
    -   Cumulative number of COVID cases per 100,000 persons
    -   Cumulative number of COVID deaths per 100,000 persons

{{< pagebreak >}}

# Additional Analyses

```{r}
#| label: fig-Figure
#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID deaths per 100,000"
#| echo: FALSE
knitr::include_graphics(here("results","figures","scatter_death_covdiff.png"))
```

```{r}
#| label: tbl-medianage_lm_model
#| tbl-cap: "Parameter estimates for median age adjusted linear regression"
#| echo: FALSE
lmtableage_betas = readRDS(here("results", "tables", "lmtableage.rds"))
knitr::kable(lmtableage_betas)
```

### 2025-04-04 code/analysis-code/statistical-analysis.R

@@ -13,6 +13,27 @@ library(patchwork)
 library(dplyr)
 library(tidyr)
 library(gt)
+library(yardstick)
+
+library(tidymodels)
+library(here)
+library(gtsummary)
+library(caret)
+library(pROC)
+library(quarto)
+library(GGally)
+library(tune)
+library(workflowsets)
+library(parsnip)
+
+library(glmnet) #LASSO
+library(ranger) #Random forest plots
+
+#Classification tree packages
+library(rpart)
+library(rpart.plot)
+
+
 #path to data
 #note the use of the here() package and not absolute paths
 data_location <- here::here("data","processed-data","project_data.rds")
@@ -21,6 +42,8 @@ data_location <- here::here("data","processed-data","project_data.rds")
 project_data <- readRDS(data_location)
 
 
+cc_project_data <- project_data %>%
+  drop_na()
 ######################################
 #Data fitting/statistical analysis
 ######################################
@@ -31,16 +54,16 @@ project_data <- readRDS(data_location)
 
 
 #Violin plots of the outcome by each categorical variable
-demo1 <- project_data %>%
+demo1 <- cc_project_data %>%
   ggplot(aes(coverage_diff, WHO_REGION)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "WHO region")
 
-demo2 <- project_data %>%
+demo2 <- cc_project_data %>%
   ggplot(aes(coverage_diff, WB_INCOME)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Country income level")
 
-demo3 <- project_data %>%
+demo3 <- cc_project_data %>%
   ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Total population (in thousands)")
 
-demo4 <- project_data %>%
+demo4 <- cc_project_data %>%
   ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Population density (persons per square km)")
 
 #Building 4 panel figure of country demographics
[…]

### 2025-04-04 products/manuscript/supplement/Supplementary-Material.qmd
Part 4 Commit 2 - Updating supplementary material and Read Me files

@@
     toc: false
-    number-sections: true
+    number-sections: false
     highlight-style: github
@@
 
-# Supplemental Information
+# Table of contents
 
-## Table of contents
+-   Project structure and content
+-   Country characteristic details
+-   Additional analyses
 
--   Code and file information
--   Country characteristic details
+# Project structure and content
 
-# Code and file information
+This project follows the following folder structure. See the `readme` files in each folder for more details.
 
-Explain here what each code/file is and does, and in which order (if any) users need to run thing to reproduce everything.
-Essentially, give a full set of instructions to re-generate everything.
+* The `assets` folder is a place holder for any files that do not belong in the data, code, results, or products folders. These items are not generated by code.
+
+* The `code` folder contains all coding files required for the project. This folder includes 3 sub-folders that do different parts of an analysis: 
+  exploratory data analysis (EDA), processing, and analysis. See the readme files in those folders for details.
+
+* The `data` folder contains all data files required for the project. There are 2 sub-folders, raw-data contains raw data files
+  downloaded from their online sourcesand processed-data which contains the cleaned and edited data sets, including a final
+  merged project dataset (project_data.rds). All files are saved as .rds files.
+
+* The `products` folder contains a subfolder for the manuscript, 
+  - The  `manuscript` subfolder contains a sub-folder for a supplementary material file.
+
+* The `results` folder contains automatically/code generated output. This includes subfolders for figures and tables. 
+  All figures are saved as .png files. Most tables are saved as .png and .rds files. Rds files can be used to reference specific 
+  values, while the png files will be displayed in the body of the manuscript since Quarto is currently unable to knit tables to 
+  Word. All content in these folders are automatically generated by code.
+
+* There are multiple special files in the repo.
+  * `readme.md`: this file contains instructions or details about the folder it
+  is located in. You are reading the project-level `README.md` file right now. There is a `readme` in almost every folder.
+  * `data-analysis-template.Rproj` is a file that tells RStudio that this is the main folder for a project. 
+  * a few "hidden" files and folders (they start with a `.` and depending on how your OS is configured, you might not see them). Those are for R/RStudio and Git/GitHub and you can ignore them.
+
+## Steps for reproducing the project
+
+To fully reproduce all steps of the project, the code should be run in the following order:
+
+1. processing-code/processing-code.qmd
+2. eda-code/eda_for_manuscript.qmd
+3. analysis-code/statistical-analysis.R
+4. products/manuscript/Manuscript.qmd
+5. products/manuscript/supplement/Supplementary-Material.qmd
+
+Note: Because all data, figures, and tables are saved in the relevant subfolders, the manuscript 
+can be produced by simply running the Manuscript.qmd file.
 

### 2025-04-16 code/eda-code/eda_for_manuscript.qmd
Part 5 Commit 1

--- 2025-02-20 code/eda-code/eda_for_manuscript.qmd
@@
 project_data <- readRDS(data_location)
+#Restricting project_data to complete rows 
+cc_project_data <- project_data %>%
+  drop_na()
 ```
@@
 ```{r}
-#histogram of differnce in coverage
-outcome_hist <- ggplot(project_data, aes(x = coverage_diff)) +
+#histogram of difference in coverage
+outcome_hist <- ggplot(cc_project_data, aes(x = coverage_diff)) +
   geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
@@
 ```{r}
-  table1 <- project_data %>% dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>% #Reordering the variables into the order I want displayed in the table
+  table1 <- cc_project_data %>% dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>% #Reordering the variables into the order I want displayed in the table
   dplyr::select(coverage_2022, coverage_2019, coverage_diff, Cumulative_case_rate, Cumulative_death_rate, HPV_NATIONAL_SCHEDULE, HPV_YEAR_INTRODUCTION, HPV_PRIM_DELIV_STRATEGY, HPV_SEX, doses_rec, WHO_REGION, WB_INCOME, total_pop,  total_pop, pop_density, median_age, pop_growth_rate, fertility_rate, death_rate, life_expect, migration) %>%
@@
   modify_header(
-    stat_0 = "**n/N (%), mean (sd), or median (IQR),** N = {n}") %>%
+    stat_0 = "**n/N (%), mean (SD), or median (IQR),** N = {n}") %>%
   modify_footnote(all_stat_cols() ~ NA) %>%
@@
     "death_rate", "life_expect", "migration")
-table1_prelim <- project_data %>%
+table1_prelim <- cc_project_data %>%
   dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
@@
   modify_header(
-    stat_0 = "n/N (%), mean (sd), or median (IQR), N = {n}",
+    stat_0 = "n/N (%), mean (SD), or median (IQR), N = {n}",
     label = "Characteristic"
   ) %>% as_tibble() %>%
-  mutate(`n/N (%), mean (sd), or median (IQR), N = 100` = if_else(is.na(`n/N (%), mean (sd), or median (IQR), N = 100`), " ", `n/N (%), mean (sd), or median (IQR), N = 100`))
+  mutate(`n/N (%), mean (SD), or median (IQR), N = 98` = if_else(is.na(`n/N (%), mean (SD), or median (IQR), N = 98`), " ", `n/N (%), mean (SD), or median (IQR), N = 98`))
 
@@
 saveRDS(table1, here("results","tables","table1.rds"))
-save_as_image(as_flex_table(table1), path = here("results","tables","table1.png"))
-as_flextable(table1_display)
+#save_as_image(as_flex_table(table1), path = #here("results","tables","table1.png"))
+#as_flextable(table1_display)
 
@@
     "Cumulative_death_rate")
-table1_cov_covid <- project_data %>%
+table1_cov_covid <- cc_project_data %>%
   dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
@@
   modify_header(
-    stat_0 = "Mean (sd) or median (IQR), N = {n}",
+    stat_0 = "Mean (SD) or median (IQR), N = {n}",
     label = "Characteristic"
   ) %>% as_tibble() %>%
-  mutate(`Mean (sd) or median (IQR), N = 100` = if_else(is.na(`Mean (sd) or median (IQR), N = 100`), " ", `Mean (sd) or median (IQR), N = 100`)) %>%
+  mutate(`Mean (SD) or median (IQR), N = 98` = if_else(is.na(`Mean (SD) or median (IQR), N = 98`), " ", `Mean (SD) or median (IQR), N = 98`)) %>%
   gt() %>%
@@
     column_labels.padding = px(2)    # Reduce padding in column headers
+  )  %>% 
+  text_case_match(
+    "coverage_2022" ~ "2022 HPV vaccination coverage",
+    "coverage_2019" ~ "2019 HPV vaccination coverage",
+    "coverage_diff" ~ "Coverage difference between 2022 and 2019",
+    "Cumulative_case_rate" ~ "Cumulative COVID cases",
+    "Cumulative_death_rate" ~ "Cumulative COVID deaths"
   )
@@
     "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "doses_rec")
-table1_vac_program <- project_data %>%
+table1_vac_program <- cc_project_data %>%
   dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
@@
   modify_header(
-    stat_0 = "n/N (%), mean (sd), or median (IQR), N = {n}",
+    stat_0 = "n/N (%), mean (SD), or median (IQR), N = {n}",
     label = "Characteristic"
   ) %>% as_tibble() %>%
-  mutate(`n/N (%), mean (sd), or median (IQR), N = 100` = if_else(is.na(`n/N (%), mean (sd), or median (IQR), N = 100`), " ", `n/N (%), mean (sd), or median (IQR), N = 100`)) %>% #Replacing NA with a space so that NA doesn't get printed in the table
+  mutate(`n/N (%), mean (SD), or median (IQR), N = 98` = if_else(is.na(`n/N (%), mean (SD), or median (IQR), N = 98`), " ", `n/N (%), mean (SD), or median (IQR), N = 98`)) %>% #Replacing NA with a space so that NA doesn't get printed in the table
   gt() %>%
@@
     column_labels.padding = px(2)    # Reduce padding in column headers
+  ) %>% 
+  text_case_match(
+    "HPV_NATIONAL_SCHEDULE" ~ "National HPV vaccination schedule",
+    "HPV_YEAR_INTRODUCTION" ~ "Year of HPV vaccine introduction",
+    "HPV_PRIM_DELIV_STRATEGY" ~ "Primary vaccine delivery strategy",
+    "HPV_SEX" ~ "Genders vaccinated",
+    "doses_rec" ~ "Number of recommended doses"
   )
@@
 ###########
-#This is creating the combined table 1 that can be used to pull numbers from 
+#This is creating the sociodemographics table for printing
 # Process the data and create a summary table using dplyr
@@
     "death_rate", "life_expect", "migration")
-table1_demo <- project_data %>%
+table1_demo <- cc_project_data %>%
   dplyr::select(-CODE, -CODE2, -Country, -year_covid) %>%
@@
     column_labels.padding = px(2)    # Reduce padding in column headers
+  ) %>% 
+  text_case_match(
+    "WHO_REGION" ~ "WHO region",
+    "AFR" ~ "African region",
+    "AMR" ~ "Region of the Americas",
+    "EMR" ~ "Eastern Mediterranean region",
+    "EUR" ~ "European region",
+    "SEAR" ~ "South-East Asian region",
+    "WPR" ~ "Western Pacific region",
+    "WB_INCOME" ~ "World Bank income level",
+    "total_pop" ~ "Total population (thousands)",
+    "pop_density" ~ "Population density (persons per square km)",
+    "median_age" ~ "Median age",
+    "pop_growth_rate" ~ "Population growth rate (%)",
+    "fertility_rate" ~ "Total fertility rate (live births/woman)",
+    "death_rate" ~ "Crude death rate (deaths per 1000 people)",
+    "life_expect" ~ "Life expectancy at birth (years)",
+    "migration" ~ "Net migration rate (per 1000 people)"
   )
@@
 
-scatter_case_covdiff <- project_data %>%
+scatter_case_covdiff <- cc_project_data %>%
   ggplot(aes(Cumulative_case_rate,coverage_diff)) + geom_point() + geom_smooth(method = lm) +
   labs(
-    x = "Number of COVID cases in 2020-2022 per 100,000 persons",
+    x = "Cumulative number of COVID cases in 2020-2022 per 100,000 persons",
     y = "Difference in HPV vaccination coverage between 2022 and 2019"
@@
 
-scatter_death_covdiff <- project_data %>%
+scatter_death_covdiff <- cc_project_data %>%
   ggplot(aes(Cumulative_death_rate, coverage_diff)) + geom_point() + geom_smooth(method = lm) +
@@
 ggsave(scatter_death_covdiff_loc, width = 8, height = 6, dpi = 300)
+```
+
+#### Creating bivariate plots for demographics and the outcome
+```{r}
+
+#Violin plots of the outcome by each categorical variable
+demo1 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, WHO_REGION)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "WHO region")
+
+demo2 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, WB_INCOME)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Country income level")
+
+demo3 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Total population (in thousands)")
+
+demo4 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Population density (persons per square km)")
+
+#Building 4 panel figure of country demographics
+bivar_demo_1 <- (demo1 | demo2) / (demo3 | demo4)
+bivar_demo_1_loc <- here("results","figures","bivar_demo_1.png")
+# Save the plot created as a PNG
+ggsave(bivar_demo_1_loc, plot = bivar_demo_1, width = 8, height = 6, dpi = 300)
+
+demo5 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, median_age)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Median age of population")
+
+demo6 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, fertility_rate)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Fertility rate (live births per woman)")
+
+demo7 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, death_rate)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Death rate (deaths per 1,000 population)")
+
+demo8 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, life_expect)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Life expectancy (in years)")
+
+demo9 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, migration)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Migration rate (per 1000 population)")
+
+#Building 5 panel figure of the rest of the demographics
+bivar_demo_2 <- (demo5 | demo6 | demo7) / (demo8 | demo9)
+bivar_demo_2_loc <- here("results","figures","bivar_demo_2.png")
+# Save the plot created as a PNG
+ggsave(bivar_demo_2_loc, plot = bivar_demo_2, width = 8, height = 6, dpi = 300)
+
+#Building 9 panel demographic figure
+demo_fig_loc <- here("results","figures","demo_fig.png")
+demo_fig <- (demo1 | demo2 | demo3) / (demo4 | demo5 | demo6) / (demo7 | demo8 | demo9) 
+ggsave(demo_fig_loc, plot = demo_fig, width = 8, height = 10, dpi = 300)
+
+#Vaccination variables
+vac1 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, HPV_NATIONAL_SCHEDULE)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "National vaccine schedule")
+
+vac2 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, HPV_PRIM_DELIV_STRATEGY)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Primary vaccination delivery strategy")
+
+vac3 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, HPV_SEX)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Genders vaccinated")
+
+vac4 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, as.factor(doses_rec))) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Number of recommended doses")
+
+vac5 <- cc_project_data %>%
+  ggplot(aes(coverage_diff, HPV_YEAR_INTRODUCTION)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Year of vaccine introduction")
+
+#Building 5 panel figure of vaccination program variables
+bivar_vac <- (vac1 | vac2 | vac3) / (vac4 | vac5)
+bivar_vac_loc <- here("results","figures","bivar_vac.png")
+# Save the plot created as a PNG
+ggsave(bivar_vac_loc, plot = bivar_vac, width = 8, height = 6, dpi = 300)
 ```
@@
 reg_tree <- rpart(coverage_diff ~ total_pop + pop_density + median_age + pop_growth_rate + fertility_rate + death_rate + life_expect + migration + Cumulative_case_rate + Cumulative_death_rate + HPV_YEAR_INTRODUCTION + WHO_REGION + WB_INCOME + HPV_NATIONAL_SCHEDULE + HPV_YEAR_INTRODUCTION + HPV_PRIM_DELIV_STRATEGY + HPV_SEX + doses_rec,
-                  data = project_data, 
+                  data = cc_project_data, 
                   method = "anova")  # anova method for regression trees

### 2025-04-16 code/processing-code/processing-code.qmd
Part 3 Commit 1
Part 5 Commit 1

@@
 #### HPV coverage and vaccination program data
+
 There are two data files to be imported for the HPV vaccination data, one file contains yearly coverage estimates and the other contains information about the HPV vaccination program.
@@
 ##### HPV vaccination coverage
+
 A total of 196 countries are included in the data set.
@@
 ##### HPV vaccination program characteristics
+
 A total of 196 countries are included in the data set.
@@
 Other characteristics (e.g., delivery strategy) appear complete.
+
 ```{r}
@@
 As a first step, I am combining the HPV vaccination coverage estimates with the HPV vaccination program information and performing some basic data checks.
-Other combinations will be performed in the _Combining all datasets_ section below.
+Other combinations will be performed in the *Combining all datasets* section below.
+
 ```{r}
@@
 #### United Nations data
+
 There are 238 countries with yearly data from 1950 to 2023.
@@
 Data will be restricted to the two years of interest during the next part of the project.
+
 ```{r}
@@
 #### COVID19 data
+
 This data source includes weekly COVD case count data on 239 countries.
@@
 Completeness of data will have to be reviewed once the data are restricted to countries with vaccination coverage estimates in the years of interest.
+
 ```{r}
@@
 ## Combining all datasets
+
 Above, HPV vaccination coverage and vaccine program characteristic data were combined. Here, I'm combining that data with the United Nations demographic characteristics and COVID burden.
@@
 
+project_data <- project_data %>%
+  mutate(delivery_strat = if_else(HPV_PRIM_DELIV_STRATEGY == "Varies by region/province", "Mixed", HPV_PRIM_DELIV_STRATEGY))
+
 #Looking at missingness for each included variable

### 2025-04-16 products/manuscript/Manuscript.qmd
Part 5 Commit 1

@@
 Human papillomavirus vaccines have been available since 2006, and as of 2019, 108 countries had introduced HPV vaccine into their national immunization programs.
-In 2020, immunization services globally experienced disruptions due to the COVID-19 pandemic.
+In 2020, immunization services, globally, experienced disruptions due to the COVID-19 pandemic.
 My objective was to assess the association between country-level COVID burden and changes in HPV vaccination coverage between 2019 and 2022.
@@
 LASSO regression and regression tree models were also considered.
-Models were compared using R^2^ values.
 
@@
 Variability in the coverage difference was highest for countries with the lowest COVID burden and decreased as COVID burden increased.
-After adjustment for all vaccination program and sociodemographic characteristics, coverage difference and COVID burden were not associated (beta = -0.000065 (0.00064), p-value = 0.919).
-COVID burden was an important predictor of difference in coverage among countries with a median resident age of \<26 years.
+After adjustment for all vaccination program and sociodemographic characteristics, coverage difference and COVID burden were not associated (𝛽= -0.000065 (0.00064), p-value = 0.919).
+Based on the regression tree analysis, COVID burden was an important predictor of difference in coverage among countries with a median resident age of \<26 years.
 
@@
 To address this gap, my objective was to assess the association between country-level COVID burden and changes in HPV vaccination coverage between 2019 and 2022.
-I hypothesized that decreases in HPV vaccination coverage between 2022 and 2019 were greater among countries with high burden of COVID-19 in 2020-2022.
+I hypothesized that decreases in HPV vaccination coverage between 2019 and 2022 would be greater among countries with high burden of COVID-19 in 2020-2022.
 A secondary objective of this work was to develop a regression tree to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccine program and sociodemographic characteristics.
-The purpose of this regression tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV coverage if a new epidemic or pandemic were to affect a country.
+The purpose of this regression tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV vaccination coverage if a new epidemic or pandemic were to affect a country.
 
@@
 Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).[@HPVdashboard] Briefly, coverage data reflect administrative and official HPV vaccination coverage reported annually through the WHO/UNICEF Joint Reporting Form on Immunization.
-The HPV Dashboard includes country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccination recommendation, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule, whether the program is gender neutral (i.e., targets males and females), WHO region, and year-specific coverage from 2010 to 2023.
+The HPV Dashboard includes country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccination recommendation, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule (1 or 2), whether the program is gender neutral (i.e., targets males and females or females only), WHO region, and year-specific coverage from 2010 to 2023.
 
@@
 COVID burden between January 4, 2020, and December 31, 2022, was defined as the cumulative number of COVID cases per 100,000 persons.
-Additionally, cumulative number of COVID deaths per 100,000 was also calculated.
+Additionally, the cumulative number of COVID deaths per 100,000 persons was also calculated.
 
@@
 For categorical variables, the distribution (n, %) of each level was calculated.
-Bivariate associations between each characteristic and the change in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots (with a LOESS line of best fit for visualization purposes only) for continuous variables and violin plots for categorical variables.
+Bivariate associations between each characteristic and the difference in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots (with a LOESS line of best fit for visualization purposes only) for continuous variables and violin plots for categorical variables.
 
@@
 Additionally, models adjusted for each country level characteristic individually were performed.
-For each model, the effect of COVID case count on the difference in HPV vaccination coverage between 2019-2022 (beta, standard error, p-value from t-test with a null hypothesis of beta = 0) and model performance metrics (R-squared, AIC, BIC) were extracted and compared.
+For each model, the effect of COVID case count on the difference in HPV vaccination coverage between 2022 and 2019 (beta, standard error, p-value from t-test with a null hypothesis of beta = 0) and model performance metrics (R-squared, AIC, BIC) were extracted and compared.
 Selection of a final model was based on subject matter expertise, which model had the highest R-squared, and which model had the lowest AIC and/or BIC.
@@
 table1_display=readRDS(here("results","tables","table1_tib.rds"))
-table1_tib <- as_tibble(table1_display) %>% rename(character = `Characteristic`, stat = `n/N (%), mean (sd), or median (IQR), N = 100`)
+table1_tib <- as_tibble(table1_display) %>% rename(character = `Characteristic`, stat = `n/N (%), mean (SD), or median (IQR), N = 98`)
 ```
@@
 
-Between January 1, 2020, and December 31, 2022, the mean (sd) number of cases per 100,000 persons was `r table1_tib %>%  filter(row_number() == which(character == "Cumulative_case_rate") + 1) %>% pull(stat)`.
+Between January 1, 2020, and December 31, 2022, the mean (sd) number of COVID cases per 100,000 persons was `r table1_tib %>%  filter(row_number() == which(character == "Cumulative_case_rate") + 1) %>% pull(stat)`.
 The median (IQR) was lower at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_case_rate") + 2) %>% pull(stat)`.(@tbl-Table1_coverage_covid) The mean 3-year COVID death rate was `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 1) %>% pull(stat)` deaths per 100,000 persons; the median was similar at `r table1_tib %>% filter(row_number() == which(character == "Cumulative_death_rate") + 2) %>% pull(stat)`.
@@
 #| label: fig-Figure2
-#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID cases per 100,000"
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID cases per 100,000"
 #| echo: FALSE
@@
 Countries with a facility-based delivery strategy, gender neutral vaccination program, and 2-dose recommendation had distributions of difference in coverage with a smaller portion of countries in the distribution region indicating a decrease in coverage (i.e., most of the data was centered around zero or was greater than 0).
-There was not a strong association with difference in coverage and year of implementation.
+There was not a strong association between difference in coverage and year of implementation.
 
@@
 #| label: fig-Figure5
-#| fig-cap: "Associations between vaccine program characteristics and difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID deaths per 100,000"
+#| fig-cap: "Associations between vaccine program characteristics and difference in HPV vaccination coverage between 2022 and 2019 and vaccination program characteristics"
 #| echo: FALSE
@@
 
-For the unadjusted and all adjusted multivariable linear regression models, COVID burden beta estimates, standard errors, p-values for the effect of cumulative COVID cases on the difference in coverage, and performance metrics (R^2^, AIC, BIC) are presented in @tbl-modelcomp.
+For the unadjusted and all adjusted multivariable linear regression models, COVID burden beta estimates, standard errors, and p-values for the effect of cumulative COVID cases on the difference in coverage as well as model performance metrics (R^2^, AIC, BIC) are presented in @tbl-modelcomp.
 Each performance metric has been shaded using a sliding color scale; a light color indicates worse performance, darker color indicates better performance, and the "best" performing model is indicated using bold text.
@@
 The model adjusted for median resident age had the lowest AIC and BIC.
-The association between change in coverage and COVID burden were similar in the fully adjusted model (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Full model")) %>% pull(beta) %>% format(scientific = FALSE)`) and model only adjusted for median age (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Median age")) %>% pull(beta) %>% format(scientific = FALSE)`) (@tbl-modelcomp, see supplementary material for all model parameters for median age adjusted model).
-Preliminary modeling will focus on the fully adjusted model.
+The association between change in coverage and COVID burden was similar in the fully adjusted model (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Full model")) %>% pull(beta) %>% format(scientific = FALSE)`) and model only adjusted for median age (𝛽=`r prelim_lm_betas %>% filter(row_number() == which(Adjustment == "Median age")) %>% pull(beta) %>% format(scientific = FALSE)`) (@tbl-modelcomp, see supplementary material for all model parameters for median age adjusted model).
+Multivariable linear regression modeling will focus on the fully adjusted model.
 However, none of these models fit particularly well based on the low R^2^ values and diagnostic plots in @fig-lm-diagnostics.
@@
 
-Because of the large number of coefficients in the full model, LASSO regression model was also performed for regularization.
+Because of the large number of coefficients in the full model, LASSO regression was also performed for regularization.
 A grid of penalty values from 1E-5 to 1E2 on a log scale was considered for tuning.
@@
 Each color represents a particular run.
-R^2^ is relatively stable for tuning parameters values from 1E-5 to 1E-2; therefore, a tuning value of 1E-2 was selected for the final model.
+R^2^ is relatively stable for tuning parameter values from 1E-5 to 1E-2; therefore, a tuning value of 1E-2 was selected for the final model.
 
@@
 There was not a significant association between cumulative number of COVID cases and the difference in HPV vaccination coverage between 2019-2022 using the linear model (𝛽(se) = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(estimate) %>% format(scientific = FALSE)` (`r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(std.error) %>% format(scientific = FALSE)`), *p*-value = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(p.value)`).
-After adjusting for all other characteristics, the difference in coverage was significantly higher and positive (indicating increase in coverage between 2019 and 2022) for all WHO regions, except the Eastern Mediterranean Region, compared to the African Region.
+After adjusting for all other characteristics, the difference in coverage was significantly higher and positive (indicating an increase in coverage between 2019 and 2022) for all WHO regions, except the Eastern Mediterranean Region, compared to the African Region.
 Conclusions from the LASSO regression model were similar.
@@
 
+{{< pagebreak >}}
 ```{r}
@@
 
-Of the four models considered, the regression tree had the largest R^2^ value (`r r_sq_table_ref %>% filter(row_number() == which(Model == "Regression tree")) %>% pull(rsq_est) %>% round(3)`).
+Of the four models considered, the regression tree had the largest R^2^ value (`r r_sq_table_ref %>% filter(row_number() == which(Model == "Regression tree")) %>% pull(rsq_est) %>% round(3)`; @tbl-rsq-models).
 This model is also potentially the most intuitive, and the utility of estimating COVID burden is apparent in this model.

### 2025-04-16 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 Commit 1

@@
 
-* The `data` folder contains all data files required for the project. There are 2 sub-folders, raw-data contains raw data files
-  downloaded from their online sourcesand processed-data which contains the cleaned and edited data sets, including a final
-  merged project dataset (project_data.rds). All files are saved as .rds files.
+* The `data` folder contains all data files required for the project. There are 2 sub-folders: raw-data contains raw data files
+  downloaded from their online sources and processed-data contains the cleaned and edited data sets, including a final
+  merged project data set (project_data.rds). All files are saved as .rds files.
 
 * The `products` folder contains a subfolder for the manuscript, 
-  - The  `manuscript` subfolder contains a sub-folder for a supplementary material file.
+  - The  `manuscript` subfolder contains a subfolder for a supplementary material file.
 
-* The `results` folder contains automatically/code generated output. This includes subfolders for figures and tables. 
+* The `results` folder contains code generated output and includes subfolders for figures and tables. 
   All figures are saved as .png files. Most tables are saved as .png and .rds files. Rds files can be used to reference specific 
@@
 * There are multiple special files in the repo.
-  * `readme.md`: this file contains instructions or details about the folder it
-  is located in. You are reading the project-level `README.md` file right now. There is a `readme` in almost every folder.
+  * `readme.md`: this file contains instructions and/or details about the folder it is located in. 
+    The information you are reading now is covered in the project-level `README.md` file. There is a `readme` in almost every folder.
   * `data-analysis-template.Rproj` is a file that tells RStudio that this is the main folder for a project. 
@@
     -   Year of HPV vaccine program introduction
-    -   Primary HPV vaccine delivery strategy (e.g., school-based)
-    -   Indicator for gender neutral vaccination (i.e., female-only vaccination or females and males)
+    -   Primary HPV vaccine delivery strategy (school-based, clinic-based, varies)
+    -   Indicator for gender neutral vaccination (female-only vaccination or females and males)
     -   Recommended number of doses (1 or 2)
@@
 #| label: fig-Figure
-#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2019 and 2022 and cumulative COVID deaths per 100,000"
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID deaths per 100,000 persons"
 #| echo: FALSE

### 2025-04-16 code/analysis-code/statistical-analysis.R

@@ -41,85 +41,15 @@ data_location <- here::here("data","processed-data","project_data.rds")
 #load data. 
 project_data <- readRDS(data_location)
 
-
+#Restricting project_data to complete rows 
 cc_project_data <- project_data %>%
   drop_na()
+
 ######################################
 #Data fitting/statistical analysis
 ######################################
 
 
-
-
-
-
-#Violin plots of the outcome by each categorical variable
-demo1 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, WHO_REGION)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "WHO region")
-
-demo2 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, WB_INCOME)) + geom_violin() + labs(x = "Difference in vaccination coverage", y = "Country income level")
-
-demo3 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, total_pop)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Total population (in thousands)")
-
-demo4 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, pop_density)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Population density (persons per square km)")
-
-#Building 4 panel figure of country demographics
-bivar_demo_1 <- (demo1 | demo2) / (demo3 | demo4)
-bivar_demo_1_loc <- here("results","figures","bivar_demo_1.png")
-# Save the plot created as a PNG
-ggsave(bivar_demo_1_loc, plot = bivar_demo_1, width = 8, height = 6, dpi = 300)
-
-
-demo5 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, median_age)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Median age of population")
-
-demo6 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, fertility_rate)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Fertility rate (live births per woman)")
-
-demo7 <- cc_project_data %>%
-  ggplot(aes(coverage_diff, death_rate)) + geom_point() + geom_smooth() + labs(x = "Difference in vaccination coverage", y = "Death rate (deaths per 1,000 population)")
-
-demo8 <- cc_project_data %>%
[…]

### 2025-04-24 project-review-Yang.qmd
Review 1 - G Yang

---
title: Project Review Overall Score - 3
author: Guozheng Yang
date: 04/21/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022

Name of project author(s): Rayleen Lewis

Name of project reviewer: Guozheng Yang

# Specific project content evaluation

## Background, Context and Motivation

### Feedback and Comments

The background is comprehensive and well-referenced, providing a clear explanation of the importance of HPV vaccination and the global impact of COVID-19 on healthcare services. Given the timing of the disruptions caused by COVID-19, this motivation is both logical and timely. The author did a great job outlining research gaps in existing literature and how these gaps motivate this study. Additionally, the author clearly stated how this study addresses the gaps by quantifying the impact of COVID-19 burden rather than relying on simple pre/post comparisons.

One minor suggestion is to compare HPV vaccination coverage rates when introducing the difference in HPV burdens among different countries. I am not sure if this information is available online, and it can be foreseen that there should be a positive correlation. But including some basic statistics will strengthen the motivation of this study.

### Summary assessment

* strong contextualization and motivation

## Question description

### Feedback and Comments

The research question is clearly stated and directly relates to the data available. The hypothesis is intuitive and well framed that higher COVID-19 burden would correlate with greater drops in HPV vaccination coverage. The secondary objective about building the tree model is also clearly described and adds on the depth of this study.

### Summary assessment

* question/hypotheses fully clear

## Data description

### Feedback and Comments

The manuscript provides detailed information about all data sources, including hyper-links and download dates. The variables are described with sufficient clarity, and the inclusion of WHO, UN, and COVID dashboards ensures credibility. Although not every variable is described in the main text, the reference to supplementary materials is appropriate. I particularly like how the author structure the README file for all data files, with each data set clearly listed. This is easy for reviewers to follow and saved me a lot of time. Also, good job with the hyper-links!

I praise the great effort the author put in cleaning the raw data.

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis

### Feedback and Comments

The data cleaning process is clearly described, with reasonable exclusions. Exploratory visualizations (histograms, scatterplots, violin plots) are well chosen and support the narrative. While alternatives for data transformation or imputation are not discussed, the decision to use complete case analysis is justified given the small number of incomplete cases. The tables list all necessary statistics for readers to understand the distribution of each variable. The code file for data processing is very well structured, with a clear logic flow and important annotations attached. The exploratory text further improves the readability of this data wrangling process.

Though the data cleaning process is pretty clear in the code file, I suggest the author to include the exclusion criteria in the main text. One or two sentences will be enough.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

### Feedback and Comments

The author conducted a comprehensive analysis not only addressing the proposed research question but also covering contents from this course. The decision to use LASSO for variable selection and to explore regression trees for interpretability demonstrates thoughtful methodology. The author also examined the residual plots for a throughout diagnosis of the fitting results, with limitations of the models acknowledged.

I wonder if the author tried generalized linear models, since the regular multivariable regression is likely limited considering the unique distribution of the response. To integrate it into the LASSO model will also be interesting. Additionally, the author may want to briefly justify the use of R-squared for model comparison instead of RMSE. One sentence or two will be good enough.

### Summary assessment

* strong and reasonable analysis

## Presentation

### Feedback and Comments

The figures and tables are clean, relevant, and well-labeled. Diagnostic plots and R-square comparisons help readers evaluate model performance. All visuals are logically placed within the manuscript.

A minor suggestion is to improve figure captions for standalone interpretability. For example, the author could include a detailed explanation of the blue lines and gray areas in captions of Figure 3 and Figure 4. It would be great if the readers can get a comprehensive understanding of the results by just looking at the figures.

### Summary assessment

* results are very well presented

## Discussion/Conclusions

### Feedback and Comments

The discussion is thoughtful and well-balanced, highlighting the implications for public health practice and the importance of data variability and methodology across countries. The author also discussed both strengths and limitations of this study, which deepens the discussion section. The conclusion is also policy-relevant and grounded in the results.

### Summary assessment

* strong, complete and clear discussion

## Further comments

Excellent work overall. The tree model is a strong highlight of the analysis and adds real-world applicability for future policy planning. One small edit: check for a typo in "gender nuetral" (should be "gender neutral").

# Overall project content evaluation

## Structure

### Feedback and Comments

The folder and file structure appear straightforward. File names are intuitive, and no unnecessary files are included. I also like the detailed instructions provided in each README file. I could easily follow the workflow in this project.

### Summary assessment

* well structured

## Documentation

### Feedback and Comments

The author retained the structural framework of the template folders. Overall, the project folder is well-organized: clean, concise, and free of redundant files. Each executable file contains a necessary README file in the same folder, which greatly enhances usability for readers. Additionally, the executable files are sufficiently annotated to explain each step, further demonstrating the author's attention to clarity and reproducibility.

### Summary assessment

* fully and well documented

## Reproducibility

### Feedback and Comments

I ran through all the code files on my end without errors. The reproducibility of this project is confirmed. The author dropped part of the code for analysis, but I don't consider it as redundant. Overall, great work!

I also saw the author inserted tables in the format of pictures. This is likely because of the rendering issue. I still give the author full credits, since I had the same issue before and I know it is hard to fix.

### Summary assessment

* fully reproducible without issues

## Thoroughness

### Feedback and Comments

The study is comprehensive and considers multiple analytic strategies. The inclusion of LASSO and tree methods strengthens the findings and offers nuanced insights beyond traditional linear modeling. Limitations are appropriately considered and well discussed. The author also listed implications for public health policy makers, which enhances the applicability of the conclusion of this study.

### Summary assessment

* strong level of thorougness

## Further comments

The project addresses an important global health issue and demonstrates a solid understanding of statistical modeling. Great use of public data and thoughtful application of advanced methods to enhance interpretability and policy relevance. The project folder is also very well organized, it is easy for me to follow each step and I believe it will be the same case for other readers.

I think the author did a great job choosing the most appropriate models to explore the research question. So I will not complain for not including comparisons across different models. I think the contents included in this project is good enough to show what the author learned from this class.

### 2025-04-25 project-review-template-Lewis-vandenHurk.qmd
Review 2 - S van den Hurk

---
title: Project Review Template 
author: Shaun van den hurk
date: 25 April 2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022

Name of project author(s): Rayleen Lewis

Name of project reviewer: Shaun van den Hurk

# Specific project content evaluation

Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation

How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

The background is very well written and there is clear contextual background in literature and practically. It is clear that the author is very well versed in the subject matter and field. The global burden of HPV, the role of vaccination, and how COVID-19 disrupted immunization services is clearly outlined. The rationale is clear as well as providing a secondary objective of policy utility for the approach provided.

### Summary assessment

-   strong contextualization and motivation

## Question description

How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The primary and secondary research questions are clearly stated and tied directly to the data. The hypothesis is testable and appropriate. The questions logically emerge from the background discussion, and the author appropriately defines the expected direction of the association.

### Summary assessment

-   question/hypotheses fully clear

## Data description

How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The author was very clear on all factors and provided very clear instructions and background. The manuscript gives a clear overview of the data sources (WHO, UN), timeframes, and variable definitions. The integration of datasets using a country identifier is explained, and relevant details (e.g., cumulative COVID burden, vaccination coverage) are well defined. 
The manuscript and description in the EDA components are already clear, but the README file in the raw-data file is particularly clear and helpful.

### Summary assessment

-   source and overall structure of data well explained

## Data wrangling and exploratory analysis

How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Very thorough, well documented and easy to follow. The data wrangling steps are described thoroughly, particularly how the outcome was calculated, how missing data were handled, and how variables were recoded. The EDA is extensive and visually supported through histograms, scatterplots, violin plots, and summary tables, with good descriptions and comments to guide this.

### Summary assessment

-   essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis

Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The project uses a strong combination of methods: univariate, multivariable linear regression, LASSO regression, and regression tree modeling. The justification for each approach is sound, and steps are explained clearly. Performance metrics (R², AIC, BIC) are used to compare models, and tuning is well documented. The author also considers overfitting and model diagnostics.

Additional code was blocked out which is not necessarily but could perhaps be included for the purpose fo the model to use the test/train split approach. But I believe that this is strong and great as a product.

### Summary assessment

-   strong and reasonable analysis

## Presentation

How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Excellent, very high quality, ready for publication. Figures and tables are well labeled, relevant, and appropriately integrated into the narrative. The regression tree is especially useful as a visual summary of subgroup findings.
Excellent use of a reproducible workflow. This is a high quality and finished product which is suitable for a high-end journal and upload in my opinion.

### Summary assessment

-   results are very well presented

## Discussion/Conclusions

Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The discussion appropriately interprets findings and acknowledges the potential for regional or methodological differences. Strengths (e.g., regression tree for policy insight) and limitations (e.g., national-level data, coverage estimate uncertainty) are clearly outlined. The conclusion is consistent with the results and provides practical insight.
The authors expertise are clearly evident.

### Summary assessment

-   strong, complete and clear discussion

## Further comments

The project is really impressive and excellent quality, and it is clear that you have a great deal of experience and expertise in this field. The project demonstrates excellent statistical reasoning and documentation.

There are very few suggestions for practical improvement.
Some very small things:
- Review the spelling in the Word document where things can be highlighted as when I viewed that document I noticed a few small spelling errors that are likely missed in Quarto (e.g nuetral on line 169). I only saw a couple and these will be quick to fix.
- Abstract: Small comment - you started by saying COVID-19 and then just said COVID, i am not sure if this was intentional or if you want to update them all to COVID-19.
- I do not know if train/test split and using CV is important for the sake of the course and if you want to add them (perhaps you did and I missed that). I do not think you need it more generally, but perhaps it is valuable to demonstrate approaches covered in the course.

# Overall project content evaluation

Evaluate overall features of the project by filling in the sections below.

## Structure

Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Everything is clearly organised and the README files were well used and outlined the approach and the code. Files and folders were well used and clealy explained. This exemplifies the concepts outlined in the course.
The only small uncertainty that I had was why there are two eda files but this was covered in the README file. This could perhaps also be added as a line at the top of bboth EDA scripts but is not necessary.

### Summary assessment

-   well structured

## Documentation

How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The methods section in the manuscript is very well documented. Variable definitions, model choices, and tuning procedures are explained in full, with logical rationale. There is good transparency around decisions, especially with model comparisons. The only suggestion could be to add some of this interpretation and outlining of the approach into the statistical analysis file script, although that is not necessary and will add to the length of the document and affect the clean look of the current document.

### Summary assessment

-   fully and well documented

## Reproducibility

Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

Excellent use of a reproducible workflow to run the full analysis. The analysis was also fast to run in its entirety. 
The only very small (tiny) intervention that was needed, was to add "library(patchwork)" to your eda_for_manuscript script so that the script would run fully. This wouldn't be an issues when you are not opening the script directly and was included in the statistical-analysis file.

### Summary assessment

-   fully reproducible without issues

## Thoroughness

How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

The author explored multiple modeling approaches, assessed diagnostics, and discussed results with nuance. Exploratory analyses were wide-ranging and the rationale for variable inclusion is clear. Overall, a very comprehensive and thoughtful analysis.
As mentioned, it is clear that the author has expertise in this field.

### Summary assessment

-   strong level of thorougness

## Further comments

Fantastic effort! This is a well-designed, well-documented, and well-executed project. I believe that it exceeds any expectations for the course and is of a very high quality that should be published in a scientific journal. It was clear to me that you have a strong background and experience in these approaches and content. The approach is very logical, clear, efficient and thorough. I learned a lot from reviewing this project and seeing the approach and application of concepts.

### 2025-04-25 code/eda-code/eda.qmd
Part 6 - Commit 1

@@
 ---
+
+# Description
+This EDA file includes all EDA conducted for the project. For EDA analyses included in the mansucript, please see the "eda_for_mansucript.qmd" file.
 
@@
 #Getting distributions for categorical variables
-freq_table <- proc_freq(project_data, tables = c("WHO_REGION","WB_INCOME","HPV_NATIONAL_SCHEDULE","HPV_YEAR_INTRODUCTION","HPV_PRIM_DELIV_STRATEGY","HPV_SEX","doses_rec"), options = "missing", output = out)
+freq_table <- procs::proc_freq(project_data, tables = c("WHO_REGION","WB_INCOME","HPV_NATIONAL_SCHEDULE","HPV_YEAR_INTRODUCTION","HPV_PRIM_DELIV_STRATEGY","HPV_SEX","doses_rec"), options = "missing", output = out)
 freqtable_file = here("results", "tables", "freq_table_results.rds")

### 2025-04-25 code/eda-code/eda_for_manuscript.qmd
Part 6 - Commit 1

@@
 All exploratory figures and tables presented in the manuscript are coded in this code file.
+For all EDA analyses conducted for this project (including those not included in the manuscript), please see the "eda.qmd" file.
 
@@
 library(kableExtra)
+library(patchwork)
+library(tidyr)
 ```

### 2025-04-25 code/processing-code/processing-code.qmd
Part 6 - Commit 1

@@
 
-#Data are 
-proc_freq(combo, c("WHO_REGION", "WB_INCOME", "HPV_NATIONAL_SCHEDULE", "HPV_YEAR_INTRODUCTION", "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "HPV_INT_DOSES"), options = "missing")
+#Data are very complete
+procs::proc_freq(combo, tables = c("WHO_REGION", "WB_INCOME", "HPV_NATIONAL_SCHEDULE", "HPV_YEAR_INTRODUCTION", "HPV_PRIM_DELIV_STRATEGY", "HPV_SEX", "HPV_INT_DOSES"), options = "missing")
 

### 2025-04-25 products/manuscript/Manuscript.qmd
Part 6 - Commit 1

@@
 ---
-title: "Effect of COVID-19 burden on changes in HPV vaccination coverage among females between 2019 and 2022"
+title: "Effect of COVID burden on changes in HPV vaccination coverage among females between 2019 and 2022"
 author: "Rayleen Lewis"
@@
 Several HPV vaccines are available to prevent infections that cause most HPV-attributable cancers.
-The first vaccine product, a quadrivalent vaccine that protects against 4 HPV types, entered the market in 2006; the United States was the first country to make a national HPV vaccination recommendation.[@markowitzTenYearsHuman2018; @bruniHPVVaccinationIntroduction2021] Since then, five additional vaccine products have entered the market and over 100 countries have introduced HPV vaccination into their national immunization program.[@whoHumanPapillomavirusVaccines2022; @bruniHPVVaccinationIntroduction2021]
+The first vaccine product, a quadrivalent vaccine that protects against 4 HPV types, entered the market in 2006; the United States was the first country to make a national HPV vaccination recommendation.[@markowitzTenYearsHuman2018; @bruniHPVVaccinationIntroduction2021] Since then, five additional vaccine products have entered the market and over 100 countries have introduced HPV vaccination into their national immunization program.
+However, many countries with the highest burden of cervical cancer have not been able to introduce HPV vaccination.[@whoHumanPapillomavirusVaccines2022; @bruniHPVVaccinationIntroduction2021]
 
@@
 To address this gap, my objective was to assess the association between country-level COVID burden and changes in HPV vaccination coverage between 2019 and 2022.
-I hypothesized that decreases in HPV vaccination coverage between 2019 and 2022 would be greater among countries with high burden of COVID-19 in 2020-2022.
-A secondary objective of this work was to develop a regression tree to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccine program and sociodemographic characteristics.
+I hypothesized that decreases in HPV vaccination coverage between 2019 and 2022 would be greater among countries with high burden of COVID in 2020-2022.
+A secondary objective of this work was to develop a regression tree to assist with classifying countries by their change in HPV vaccination coverage using other country-level HPV vaccination program and sociodemographic characteristics.
 The purpose of this regression tree is to identify key sociodemographic indicators that may be used by policy makers for pandemic preparedness efforts to prevent reductions in HPV vaccination coverage if a new epidemic or pandemic were to affect a country.
@@
 Data on HPV vaccination coverage and vaccination program details are available at the country level through the World Health Organization's (WHO) [HPV Dashboard](https://www.who.int/teams/immunization-vaccines-and-biologicals/diseases/human-papillomavirus-vaccines-(HPV)/hpv-clearing-house/hpv-dashboard).[@HPVdashboard] Briefly, coverage data reflect administrative and official HPV vaccination coverage reported annually through the WHO/UNICEF Joint Reporting Form on Immunization.
-The HPV Dashboard includes country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccination recommendation, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule (1 or 2), whether the program is gender neutral (i.e., targets males and females or females only), WHO region, and year-specific coverage from 2010 to 2023.
+The HPV Dashboard includes country income level (low, lower middle, upper middle, high), whether there was a national HPV vaccination recommendation, year of vaccine introduction, primary delivery strategy (school-based, facility based, varies by region, mixed), number of doses in routine schedule (1 or 2), whether the program is gender neutral (i.e., targets males and females or females only), WHO region (African Region (AFR), Region of the Americas (AMR), South-East Asia Region (SEAR), European Region (EUR), Eastern Mediterranean Region (EMR), and Western Pacific Region (WPR)), and year-specific coverage from 2010 to 2023.
 
-Country-specific weekly COVID-19 case and death counts are available starting on January 4, 2020, through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o).[@COVID19Dashboard] In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID-19 data.
+Country-specific weekly COVID case and death counts are available starting on January 4, 2020, through the WHO [COVID-19 dashboard](https://data.who.int/dashboards/covid19/data?n=o).[@COVID19Dashboard] In August 2023, WHO stopped requiring weekly reporting, although some countries continued to voluntarily report COVID data.
 
@@
 The outcome variable was coded as HPV vaccination coverage of at least one dose among females in 2022 minus coverage in 2019.
-Of the 100 countries with coverage information from both years, 2 countries were missing some vaccine program characteristic information, one of which was also missing income level.
+Of the 100 countries with coverage information from both years, 2 countries were missing some vaccination program characteristic information, one of which was also missing income level.
 All other countries had complete data for all variables.
+Due to the low number of countries in the analysis and high diversity among countries on demographic characteristics, I prioritized using all data in modeling rather than using a test/train split or cross validation.
 Complete cases were used for all analyses (N = 98).
@@
 COVID burden between January 4, 2020, and December 31, 2022, was defined as the cumulative number of COVID cases per 100,000 persons.
-Additionally, the cumulative number of COVID deaths per 100,000 persons was also calculated.
+Additionally, the cumulative number of COVID deaths per 100,000 persons was calculated.
 
@@
 For categorical variables, the distribution (n, %) of each level was calculated.
-Bivariate associations between each characteristic and the difference in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots (with a LOESS line of best fit for visualization purposes only) for continuous variables and violin plots for categorical variables.
+Bivariate associations between each characteristic and the difference in HPV vaccination coverage between 2022 and 2019 were visually assessed using scatterplots (with a locally estimated scatterplot smoothing (LOESS) line of best fit for visualization purposes only) for continuous variables and violin plots for categorical variables.
 
@@
 Additionally, models adjusted for each country level characteristic individually were performed.
-For each model, the effect of COVID case count on the difference in HPV vaccination coverage between 2022 and 2019 (beta, standard error, p-value from t-test with a null hypothesis of beta = 0) and model performance metrics (R-squared, AIC, BIC) were extracted and compared.
-Selection of a final model was based on subject matter expertise, which model had the highest R-squared, and which model had the lowest AIC and/or BIC.
+For each model, the effect of COVID case count on the difference in HPV vaccination coverage between 2022 and 2019 (𝛽, standard error, p-value from t-test with a null hypothesis of 𝛽 = 0) and model performance metrics (R^2^, AIC, BIC) were extracted and compared.
+Selection of a final model was based on subject matter expertise, which model had the highest R^2^, and which model had the lowest AIC and/or BIC.
 Model parameters were presented for the final selected model.
@@
 R^2^ was estimated for the best fitting model for comparison with the multivariable linear and LASSO regressions.
-
-{{< pagebreak >}}
 
@@
 knitr::include_graphics(here("results","tables","table1_cov_covid.png"))
-#knitr::knit_print(as_flex_table(table1_display))
-#table1_df=readRDS(here("results","tables","table1_df.rds"))
-#as_flextable(table1_df)
-#table1_df=readRDS(here("results","tables","table1_df.rds"))
-#knitr::knit_print(table1_df)
-#doc <- read_docx(here("products","manuscript","manuscript.docx"))
 
-# Add the gt table to the Word document
-#doc <- doc %>%
-#  body_add_gt(gt_table = table1, align = "center")
-
-# Save the Word document
-#print(doc, target = "word_table.docx")
-
-#knitr::kable(
-#  as_kable(table1_display))
-
-#table1_tbl_sum=readRDS(here("results","tables","table1_tbl_sum.rds"))
-
-#table1_tbl=readRDS(here("results","tables","try.rds"))
-#table1_tbl
 ```
@@
 The median year of vaccine introduction was `r table1_tib %>% filter(row_number() == which(character == "HPV_YEAR_INTRODUCTION") + 2) %>% pull(stat)`.(@tbl-table1_vac_program) School-based delivery programs were the most common (`r table1_tib %>% filter(row_number() == which(character == "School-based")) %>% pull(stat)`).
-A majority of programs had a gender nuetral vaccination recommendation and a 2-dose routine recommendation (rather than a 1-dose recommendation).
+A majority of programs had a gender neutral vaccination recommendation and a 2-dose routine recommendation (rather than a 1-dose recommendation).
 
@@
 There was a weak linear association between cumulative COVID cases and the difference in HPV vaccination coverage between 2019 and 2022 (@fig-Figure2).
-Of note, there is a lot of variation in the coverage difference among countries with fewer cumulative COVID cases per 100,000 persons.
+Of note, there was a lot of variation in the coverage difference among countries with fewer cumulative COVID cases per 100,000 persons.
 
@@
 #| label: fig-Figure2
-#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID cases per 100,000"
+#| fig-cap: "Scatterplot of difference in HPV vaccination coverage between 2022 and 2019 and cumulative COVID cases per 100,000.
+Black dots denote observed data; the blue line denotes the modeled linear association; the dark gray band denotes the 95% confidence interval around the modeled line of best fit."
 #| echo: FALSE
@@
 The distribution of the difference in vaccination coverage by each country-level demographic variable can be seen in @fig-Figure4.
-For each WHO region, the distribution of difference in coverage centers around 0, with the SEAR and EUR showing a more narrow distribution.
+For each WHO region, the distribution of difference in coverage centered around 0, with the SEAR and EUR showing a more narrow distribution.
 Similarly, the distribution of coverage difference was centered around 0 regardless of country income level, and high-income countries had the tightest distribution.
-The patterns in the continuous variables are less straightforward, particularly as there appear to be potential outliers.
-Based on the figure below, total population may have a linear, positive association with coverage difference, while life expectancy and median age show a negative quadratic association and fertility rate may show a positive quadratic association.
+The patterns in the continuous variables were less straightforward, particularly as there appear to be potential outliers.
+Based on the figure below, total population may have had a linear, positive association with coverage difference, while life expectancy and median age showed a negative quadratic association and fertility rate may have had a positive quadratic association.
 
@@
 #| label: fig-Figure4
-#| fig-cap: "Distribution of difference in HPV vaccination coverage between 2022 and 2019 by each country demographic variable"
+#| fig-cap: "Distribution of difference in HPV vaccination coverage between 2022 and 2019 by each country demographic variable.
+For scatterplots, black dots denote observed data; the blue line denotes the modeled line of best fit using LOESS; the dark gray band denotes the 95% confidence interval around the modeled line of best fit."
 #| echo: FALSE
@@
 
-There were some differences in the distribution of coverage difference by vaccine program characteristics (@fig-Figure5).
+There were some differences in the distribution of coverage difference by vaccination program characteristics (@fig-Figure5).
 Countries with a facility-based delivery strategy, gender neutral vaccination program, and 2-dose recommendation had distributions of difference in coverage with a smaller portion of countries in the distribution region indicating a decrease in coverage (i.e., most of the data was centered around zero or was greater than 0).
@@
 #| label: fig-Figure5
-#| fig-cap: "Associations between vaccine program characteristics and difference in HPV vaccination coverage between 2022 and 2019 and vaccination program characteristics"
+#| fig-cap: "Associations between vaccination program characteristics and difference in HPV vaccination coverage between 2022 and 2019 and vaccination program characteristics.
+For scatterplots, black dots denote observed data; the blue line denotes the modeled line of best fit using LOESS; the dark gray band denotes the 95% confidence interval around the modeled line of best fit."
 #| echo: FALSE
@@
 prelim_lm_betas <- tibble(prelim_lm_betas[[1]][[1]])
-#x <- prelim_lm_betas[[1]][[1]]
+
 ```
@@
 However, none of these models fit particularly well based on the low R^2^ values and diagnostic plots in @fig-lm-diagnostics.
-Based on the predicted vs observed plot, many observations with relatively high or low observed coverage difference values are further from the line of perfect fit.
-Additionally, there appears to be a slight pattern in the residuals; relatively small predicted values are more likely to have negative residuals and moderate values are more likely to have large positive values.
-Given that the beta is switching directions (positive/negative) based on the adjustment variables and is very close to zero (@tbl-modelcomp), there does not appear to be a strong association between COVID burden and difference in coverage regardless of adjustment.
+Based on the predicted vs observed plot, many observations with relatively high or low observed coverage difference values were further from the line of perfect fit.
+Additionally, there appeared to be a slight pattern in the residuals; relatively small predicted values were more likely to have negative residuals and moderate values were more likely to have large positive values.
+Given that the beta switched directions (positive/negative) based on the adjustment variables and was very close to zero (@tbl-modelcomp), there did not appear to be a strong association between COVID burden and difference in coverage regardless of adjustment.
 
@@
 A grid of penalty values from 1E-5 to 1E2 on a log scale was considered for tuning.
-@Fig-Lasso_fit shows the R^2^ value for each model tuning parameter in the left panel, and the residuals from the run with each tuning parameter in the right panel.
+@Fig-Lasso_fit shows the R^2^ value for each model tuning parameter in the left panel and the residuals from the run with each tuning parameter in the right panel.
 Each color represents a particular run.
-R^2^ is relatively stable for tuning parameter values from 1E-5 to 1E-2; therefore, a tuning value of 1E-2 was selected for the final model.
+R^2^ was relatively stable for tuning parameter values from 1E-5 to 1E-2; therefore, a tuning value of 1E-2 was selected for the final model.
 
@@
 #| label: fig-Lasso_fit
-#| fig-cap: "LASSO regression fit diagnostic plots showing the effect of tuning parameter on model R^2^ (left panel) and residuals by predicted values for each tuning parameter (right panel)"
+#| fig-cap: "LASSO regression fit diagnostic plots showing the effect of tuning parameter on model R^2^ (left panel) and residuals by predicted values for each tuning parameter (right panel) with each color corresponding to a run with a given tuning paramter value"
 #| echo: FALSE
@@
 Estimates from the fully adjusted model are shown in @tbl-fully_adj.
-There was not a significant association between cumulative number of COVID cases and the difference in HPV vaccination coverage between 2019-2022 using the linear model (𝛽(se) = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(estimate) %>% format(scientific = FALSE)` (`r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(std.error) %>% format(scientific = FALSE)`), *p*-value = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(p.value)`).
+There was not a significant association between cumulative number of COVID cases and the difference in HPV vaccination coverage between 2019-2022 using the linear model after adjustment for vaccination program and demographic characteristics (𝛽(se) = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(estimate) %>% format(scientific = FALSE)` (`r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(std.error) %>% format(scientific = FALSE)`), *p*-value = `r reg_betas %>% filter(row_number() == which(term == "Cumulative_case_rate")) %>% pull(p.value)`).
 After adjusting for all other characteristics, the difference in coverage was significantly higher and positive (indicating an increase in coverage between 2019 and 2022) for all WHO regions, except the Eastern Mediterranean Region, compared to the African Region.
@@
 
-{{< pagebreak >}}
 ```{r}
@@
 
-A strength of this work is the development of the regression tree analysis that may be used by countries if faced with a pandemic in the future.
+A strength of this work is the development of the regression tree that may be used by countries if faced with a pandemic in the future.
 Among countries with a younger median age of its residents, countries who are experiencing a higher burden of disease due to a pandemic may choose to prioritize  resources to minimize changes in vaccination coverage.

### 2025-04-25 code/analysis-code/statistical-analysis.R

@@ -582,152 +582,3 @@ r_sq_table <- tibble(Model = c("Full multivariable linear regression","Median ag
 r_sq_table_loc = here("results", "tables", "r_sq_table.rds")
 saveRDS(r_sq_table, file = r_sq_table_loc)
 
-##### Below is code that may or may not be added based on feedback from the project reviews
-# 
-# plibrary(randomForest)
-# install.packages("randomForest")
-# 
-# # Fit a Random Forest model
-# rf_model <- randomForest(coverage_diff ~ total_pop + pop_density + median_age + pop_growth_rate + 
-#                            fertility_rate + death_rate + life_expect + migration + Cumulative_case_rate + 
-#                            HPV_YEAR_INTRODUCTION + WHO_REGION + WB_INCOME + HPV_NATIONAL_SCHEDULE + 
-#                            HPV_PRIM_DELIV_STRATEGY + HPV_SEX + doses_rec, 
-#                          data = cc_project_data, 
-#                          ntree = 500,        # Number of trees (default is 500)
-#                          mtry = 4,           # Number of predictors randomly selected at each split
-#                          importance = TRUE,  # Compute variable importance
-#                          na.action = na.omit)  # Remove missing values
-# 
-# # View model summary
-# print(rf_model)
-# 
-# # Variable importance plot
-# varImpPlot(rf_model)
-
-# Leaving the code below here in case I want to add a table of mean coverage diff and correlations by demographics
-
-# 
-# # Define categorical and continuous variables
-# categorical_vars <- c("WHO_REGION", "WB_INCOME", "HPV_NATIONAL_SCHEDULE", 
-#                       "HPV_YEAR_INTRODUCTION", "HPV_PRIM_DELIV_STRATEGY", 
-#                       "HPV_SEX", "doses_rec")
-# 
-# continuous_vars <- c("Cumulative_case_rate", "Cumulative_death_rate", 
-#                      "total_pop", "pop_density", "median_age", 
-#                      "pop_growth_rate", "fertility_rate", "death_rate", 
-#                      "life_expect", "migration")
-# 
-# # Assuming `cc_project_data` contains all variables
-# # Function to compute Mean (SD) and ANOVA p-value for categorical variables
-# categorical_summary <- categorical_vars %>%
-#   lapply(function(var) {
-#     # Run ANOVA
[…]
